---
title: "All Countries"
format: html
page-layout: full
---

```{r}
#| label: set up 
#| include: false
#| warning: false

source("V1/Scripts/V1_Dashboard_setup.R")
message("✅ _setup.R successfully loaded!")

# Prepare all countries data frame
# Get unique countries with their metadata
country_name_col <- dplyr::case_when(
  "country" %in% names(country_summary_df) ~ "country",
  "Country" %in% names(country_summary_df) ~ "Country",
  TRUE ~ NA_character_
)

if (!is.na(country_name_col)) {
  # Get latest data for each country
  all_countries_data <- country_summary_df %>%
    dplyr::mutate(country_label = .data[[country_name_col]]) %>%
    dplyr::filter(
      !is.na(country_label) & country_label != "",
      !is.na(Month),
      Month <= recent_month
    ) %>%
    dplyr::group_by(country_label) %>%
    dplyr::arrange(Month, .by_group = TRUE) %>%
    dplyr::slice_tail(n = 1) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      region_label = dplyr::coalesce(.data[["Region"]], ""),
      region_slug = if_else(
        nzchar(region_label),
        tolower(gsub(" & |, | ", "", region_label)),
        NA_character_
      ),
      region_href = if_else(
        !is.na(region_slug),
        paste0(region_slug, ".html"),
        NA_character_
      ),
      # Generate blurb text similar to severity countries
      severity_blurb = severity_country_blurb(
        country_label,
        cum_ratio,
        cum_high,
        region = region_label,
        region_href = region_href
      ),
      # Ensure we have plot available
      has_plot = country_label %in% names(all_country_plots)
    ) %>%
    # Only keep countries that have plots
    dplyr::filter(has_plot) %>%
    dplyr::arrange(country_label)
  
  # Merge data status info with all_countries_data (reuse from setup script)
  if (exists("country_data_status") && nrow(country_data_status) > 0) {
    all_countries_data <- all_countries_data %>%
      dplyr::left_join(
        country_data_status,
        by = c("country_label" = "country_name")
      ) %>%
      dplyr::mutate(
        has_estimated = dplyr::if_else(is.na(has_estimated), FALSE, has_estimated),
        has_observed = dplyr::if_else(is.na(has_observed), FALSE, has_observed),
        data_status_message = dplyr::if_else(
          is.na(data_status_message),
          "Recent months contain observed data only",
          data_status_message
        )
      )
  } else {
    # Fallback if country_data_status doesn't exist
    all_countries_data <- all_countries_data %>%
      dplyr::mutate(
        has_estimated = FALSE,
        has_observed = TRUE,
        data_status_message = "Recent months contain observed data only"
      )
  }
} else {
  all_countries_data <- dplyr::tibble(
    country_label = character(),
    region_label = character(),
    cum_ratio = numeric(),
    cum_high = numeric(),
    severity_blurb = character(),
    region_href = character(),
    has_plot = logical()
  )
}

```

```{r}
#| label: last-update
#| include: false

if (!exists("latest_update_label") || !exists("latest_nowcast_dir")) {
  latest_dataset_info <- get_latest_dataset_info()
  latest_update_label <- latest_dataset_info$label
  latest_nowcast_dir <- latest_dataset_info$dir
}

```

<div class="global-summary-callout">
  <span class="global-summary-label">Data last updated</span>
  <span class="global-summary-date">`r latest_update_label`</span>
</div>

Explore all countries with available dengue surveillance data. Use the search and sort controls below to filter by name, region, cumulative cases, or severity ratio.

```{r}
#| label: search-sort-controls
#| echo: false
#| results: 'asis'

# Create search and sort controls using htmltools
search_controls <- htmltools::div(
  class = "country-search-controls",
  style = "margin: 2rem 0; max-width: 1200px; margin-left: auto; margin-right: auto;",
  htmltools::div(
    style = "display: flex; flex-wrap: wrap; gap: 1rem; align-items: center; justify-content: center;",
    htmltools::div(
      style = "flex: 1; min-width: 250px;",
      htmltools::tags$label(
        `for` = "country-search",
        style = "display: block; margin-bottom: 0.5rem; font-weight: 600;",
        "Search by name:"
      ),
      htmltools::tags$input(
        type = "text",
        id = "country-search",
        placeholder = "Type country name...",
        style = "width: 100%; padding: 0.6rem; border: 1px solid #ddd; border-radius: 6px; font-size: 1rem;"
      )
    ),
    htmltools::div(
      style = "flex: 1; min-width: 200px;",
      htmltools::tags$label(
        `for` = "region-filter",
        style = "display: block; margin-bottom: 0.5rem; font-weight: 600;",
        "Filter by region:"
      ),
      htmltools::tags$select(
        id = "region-filter",
        style = "width: 100%; padding: 0.6rem; border: 1px solid #ddd; border-radius: 6px; font-size: 1rem;",
        htmltools::tags$option(value = "", "All regions")
      )
    ),
    htmltools::div(
      style = "flex: 1; min-width: 200px;",
      htmltools::tags$label(
        `for` = "sort-by",
        style = "display: block; margin-bottom: 0.5rem; font-weight: 600;",
        "Sort by:"
      ),
      htmltools::tags$select(
        id = "sort-by",
        style = "width: 100%; padding: 0.6rem; border: 1px solid #ddd; border-radius: 6px; font-size: 1rem;",
        htmltools::tags$option(value = "name", "Name (A-Z)"),
        htmltools::tags$option(value = "name-desc", "Name (Z-A)"),
        htmltools::tags$option(value = "region", "Region"),
        htmltools::tags$option(value = "cases", "Cumulative Cases (High to Low)"),
        htmltools::tags$option(value = "cases-asc", "Cumulative Cases (Low to High)"),
        htmltools::tags$option(value = "ratio", "Severity Ratio (High to Low)"),
        htmltools::tags$option(value = "ratio-asc", "Severity Ratio (Low to High)")
      )
    )
  )
)

# Output the HTML properly for Quarto
knitr::asis_output(as.character(search_controls))

```

```{r}
#| label: all-countries-grid
#| echo: false
#| message: false
#| warning: false

if (nrow(all_countries_data) == 0) {
  panel <- htmltools::div(
    class = "alert alert-info",
    "Country data is still loading. Please check back soon."
  )
} else {
  # Create cards for all countries
  country_cards <- purrr::map(seq_len(nrow(all_countries_data)), function(idx) {
    row <- all_countries_data[idx, ]
    country_name <- row$country_label[[1]]
    blurb_text <- row$severity_blurb[[1]]
    region_name <- row$region_label[[1]]
    region_href <- row$region_href[[1]]
    cum_cases <- row$cum_high[[1]]
    cum_ratio <- row$cum_ratio[[1]]
    plot_obj <- all_country_plots[[country_name]]
    
    plot_block <- if (is.null(plot_obj)) {
      htmltools::div(
        class = "plot-placeholder",
        "Plot not yet available for this country."
      )
    } else {
      htmltools::plotTag(plot_obj, alt = paste(country_name, "yearly outlook"))
    }
    
    # Create region link if href is available (using same format as high severity countries)
    region_display <- if (!is.na(region_href) && region_href != "") {
      htmltools::div(
        style = "text-align: center; margin-bottom: 0.5rem;",
        htmltools::tags$a(
          href = region_href,
          class = "severity-region-link",
          region_name
        )
      )
    } else {
      htmltools::tags$small(
        class = "text-muted",
        style = "display: block; text-align: center; margin-bottom: 0.5rem; font-size: 0.85rem;",
        region_name
      )
    }
    
    # Add data attributes for filtering/sorting
    htmltools::div(
      class = "snapshot-card severity-card country-card",
      `data-country-name` = tolower(country_name),
      `data-region` = tolower(region_name),
      `data-cum-cases` = ifelse(is.na(cum_cases), 0, cum_cases),
      `data-cum-ratio` = ifelse(is.na(cum_ratio), 0, cum_ratio),
      htmltools::tags$h4(country_name, class = "text-center fw-semibold"),
      region_display,
      htmltools::div(class = "severity-plot", plot_block),
      htmltools::tags$p(
        class = "severity-blurb",
        htmltools::HTML(blurb_text)
      ),
      htmltools::tags$small(
        class = "severity-footnote text-muted",
        row$data_status_message[[1]]
      )
    )
  })
  
  panel <- htmltools::div(
    class = "snapshot-grid severity-grid",
    id = "country-grid",
    country_cards
  )
}

htmltools::tagList(panel)

```

```{r}
#| label: populate-region-filter
#| echo: false
#| results: 'asis'

# Store regions as a JavaScript array for the script to use
if (nrow(all_countries_data) > 0) {
  unique_regions <- sort(unique(all_countries_data$region_label))
  # Create JavaScript array string manually (no dependency on jsonlite)
  region_json <- paste0('["', paste(unique_regions, collapse = '","'), '"]')
  cat(sprintf('<script>window.countryRegions = %s;</script>', region_json))
} else {
  cat('<script>window.countryRegions = [];</script>')
}

```

<script>
(function() {
  // Wait for DOM to be ready
  function initCountryExplorer() {
    // Get DOM elements
    const searchInput = document.getElementById('country-search');
    const regionFilter = document.getElementById('region-filter');
    const sortSelect = document.getElementById('sort-by');
    const grid = document.getElementById('country-grid');
    
    if (!grid) {
      console.warn('Country grid not found');
      return;
    }
    
    // Populate region filter dropdown
    if (regionFilter && window.countryRegions) {
      window.countryRegions.forEach(region => {
        const option = document.createElement('option');
        option.value = region.toLowerCase();
        option.textContent = region;
        regionFilter.appendChild(option);
      });
    }
    
    const cards = Array.from(grid.querySelectorAll('.country-card'));
    
    if (cards.length === 0) {
      console.warn('No country cards found');
      return;
    }
    
    // Filter and sort function
    function filterAndSort() {
    const searchTerm = (searchInput?.value || '').toLowerCase().trim();
    const regionTerm = (regionFilter?.value || '').toLowerCase().trim();
    const sortValue = sortSelect?.value || 'name';
    
    // Filter cards
    let visibleCards = cards.filter(card => {
      const countryName = card.getAttribute('data-country-name') || '';
      const region = card.getAttribute('data-region') || '';
      
      const matchesSearch = !searchTerm || countryName.includes(searchTerm);
      const matchesRegion = !regionTerm || region === regionTerm;
      
      return matchesSearch && matchesRegion;
    });
    
    // Sort cards
    visibleCards.sort((a, b) => {
      const nameA = (a.getAttribute('data-country-name') || '').toLowerCase();
      const nameB = (b.getAttribute('data-country-name') || '').toLowerCase();
      const regionA = (a.getAttribute('data-region') || '').toLowerCase();
      const regionB = (b.getAttribute('data-region') || '').toLowerCase();
      const casesA = parseFloat(a.getAttribute('data-cum-cases') || 0);
      const casesB = parseFloat(b.getAttribute('data-cum-cases') || 0);
      const ratioA = parseFloat(a.getAttribute('data-cum-ratio') || 0);
      const ratioB = parseFloat(b.getAttribute('data-cum-ratio') || 0);
      
      switch(sortValue) {
        case 'name':
          return nameA.localeCompare(nameB);
        case 'name-desc':
          return nameB.localeCompare(nameA);
        case 'region':
          if (regionA !== regionB) return regionA.localeCompare(regionB);
          return nameA.localeCompare(nameB);
        case 'cases':
          if (casesB !== casesA) return casesB - casesA;
          return nameA.localeCompare(nameB);
        case 'cases-asc':
          if (casesA !== casesB) return casesA - casesB;
          return nameA.localeCompare(nameB);
        case 'ratio':
          if (ratioB !== ratioA) return ratioB - ratioA;
          return nameA.localeCompare(nameB);
        case 'ratio-asc':
          if (ratioA !== ratioB) return ratioA - ratioB;
          return nameA.localeCompare(nameB);
        default:
          return nameA.localeCompare(nameB);
      }
    });
    
    // Hide all cards first
    cards.forEach(card => {
      card.style.display = 'none';
    });
    
    // Show and reorder visible cards
    visibleCards.forEach(card => {
      card.style.display = '';
      grid.appendChild(card);
    });
    
    // Show message if no results
    let noResultsMsg = document.getElementById('no-results-message');
    if (visibleCards.length === 0) {
      if (!noResultsMsg) {
        noResultsMsg = document.createElement('div');
        noResultsMsg.id = 'no-results-message';
        noResultsMsg.className = 'alert alert-warning text-center';
        noResultsMsg.style.margin = '2rem auto';
        noResultsMsg.style.maxWidth = '600px';
        noResultsMsg.textContent = 'No countries match your search criteria.';
        grid.parentNode.insertBefore(noResultsMsg, grid);
      }
      noResultsMsg.style.display = 'block';
      grid.style.display = 'none';
    } else {
      if (noResultsMsg) noResultsMsg.style.display = 'none';
      grid.style.display = 'grid';
    }
    }
    
    // Add event listeners
    if (searchInput) {
      searchInput.addEventListener('input', filterAndSort);
    }
    if (regionFilter) {
      regionFilter.addEventListener('change', filterAndSort);
    }
    if (sortSelect) {
      sortSelect.addEventListener('change', filterAndSort);
    }
    
    // Initial sort
    filterAndSort();
  }
  
  // Initialize when DOM is ready
  if (document.readyState === 'loading') {
    document.addEventListener('DOMContentLoaded', initCountryExplorer);
  } else {
    // DOM is already ready
    initCountryExplorer();
  }
})();
</script>

