---
title: "GDO Forecasting — Evaluation Notebook"
subtitle: "Results, figures and working notes for the forecasting objective"
author: "K M Susong"
date: today
format:
  html:
    grid:
      body-width: 1000px
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(lubridate)
library(ggplot2)
library(knitr)

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 4.5, dpi = 150
)

theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold")))

# --- Locate the repo root, whatever directory this renders from ---
find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/")
  repeat {
    if (file.exists(file.path(d, "Scripts", "V1_Pipeline.R"))) return(d)
    parent <- dirname(d)
    if (identical(parent, d)) {
      stop("DENV_global_observatory repo root not found from ", start)
    }
    d <- parent
  }
}
repo_root <- find_repo_root()

# Pinned parameters and provisional thresholds come straight from the pipeline
# config, so this notebook always reports what the scripts actually used.
source(file.path(repo_root, "Scripts", "forecasting", "00_config.R"), local = TRUE)

fc_out <- file.path(repo_root, forecast_out)

# Load a forecasting output if it exists yet, else NULL.
read_fc <- function(...) {
  p <- file.path(fc_out, ...)
  if (!file.exists(p)) return(NULL)
  readr::read_csv(p, show_col_types = FALSE)
}

# Emit a "not generated yet" callout for sections whose script has not run.
# asis_output renders correctly whether or not the calling chunk is `asis`.
pending <- function(script) {
  knitr::asis_output(sprintf(
    "::: {.callout-note appearance=\"simple\"}\nNot generated yet -- run `%s` and re-render.\n:::\n",
    script
  ))
}
```

```{r}
#| label: load-outputs

panel          <- read_fc("training_data", paste0("training_panel_", snapshot_date, ".csv"))
region_lookup  <- read_fc("training_data", "country_region_lookup.csv")
summary_country <- read_fc("data_summary", "summary_country.csv")
summary_region  <- read_fc("data_summary", "summary_region.csv")
strata_country  <- read_fc("data_summary", "strata_country.csv")
coverage_grid   <- read_fc("cv_splits", "rolling_origins.csv")
stage0_diag     <- read_fc("stage0_fit", "diagnostics.csv")

stage1_scores_row        <- read_fc("stage1_hindcast", "scores", "scores_row.csv")
stage1_scores_by_lead    <- read_fc("stage1_hindcast", "scores", "scores_by_lead.csv")
stage1_scores_by_stratum <- read_fc("stage1_hindcast", "scores", "scores_by_stratum.csv")
stage1_leaderboard       <- read_fc("stage1_hindcast", "scores", "leaderboard.csv")
stage1_pit               <- read_fc("stage1_hindcast", "scores", "pit.csv")
stage1_coverage          <- read_fc("stage1_hindcast", "scores", "coverage.csv")
stage1_operational       <- read_fc("stage1_hindcast", "scores", "operational.csv")

if (!is.null(panel)) {
  panel <- panel %>%
    mutate(
      date = as.Date(date),
      data_status = factor(data_status,
        levels = c("observed", "corrected", "estimated", "gap"))
    )
}
```

## About this notebook {.unnumbered}

Results record for the GDO forecasting objective. Not the task log (Logseq) and not pipeline code (`Scripts/forecasting/`) — this notebook only reads its outputs.

::: {.callout-note appearance="simple"}
Pinned snapshot: `Output/`r snapshot_date`/`. Re-pin in `00_config.R` and re-run the pipeline to refresh.
:::

Plan of record: `planning/2026_forecast_evaluation/03_Evaluation_Plan_DRAFT`. Stages: **0** in-sample fit · **1** rolling-origin hindcast (5/7/10 yr + expanding) · **2** prospective shadow deployment.

------------------------------------------------------------------------

# Training data {#sec-training-data}

`01_prepare_training_data.R` merges the deep historical record (`full_data_season_monthly_proportions.csv`, complete seasons) with the current-period nowcast output (`DENV_cases_nowcast_output.csv`). The nowcast view wins on any overlapping country-month, so the panel runs as far forward as data exists and carries the post-nowcast "ground truth".

`data_status` is preserved throughout: **observed** (raw report), **corrected** (reporting-delay factor applied), **estimated** (seasonal-proportion nowcast), **gap** (internal month with no value — the historical file drops all-zero years).

```{r}
#| label: status-palette

status_pal <- c(
  observed  = "#4C72B0",
  corrected = "#DD8452",
  estimated = "#C44E52",
  gap       = "grey85"
)

if (!is.null(panel)) {
  panel <- panel %>%
    mutate(region = factor(region, levels = region_levels))
}
```

```{r}
#| label: tbl-panel-overview
#| tbl-cap: "Training panel: coverage and composition."

if (is.null(panel)) {
  pending("Scripts/forecasting/01_prepare_training_data.R")
} else {
  fmt <- function(x) formatC(x, big.mark = ",", format = "d")
  tibble::tibble(
    Metric = c("Countries", "Country-months", "Date range",
               "Observed", "Corrected (reporting-delay factor)",
               "Estimated (seasonal-proportion nowcast)",
               "Gap (internal month, no value)"),
    Value = c(
      fmt(n_distinct(panel$iso3)),
      fmt(nrow(panel)),
      paste(format(min(panel$date)), "to", format(max(panel$date))),
      fmt(sum(panel$data_status == "observed")),
      fmt(sum(panel$data_status == "corrected")),
      fmt(sum(panel$data_status == "estimated")),
      fmt(sum(panel$data_status == "gap"))
    )
  ) %>% knitr::kable(align = c("l", "r"))
}
```

```{r}
#| label: fig-data-status-year
#| fig-cap: "Data-status composition by calendar year. Recent years carry the reporting-delay corrections and seasonal-proportion estimates the forecast layer treats as ground truth."
#| fig-height: 3.6

if (!is.null(panel)) {
  panel %>%
    filter(Year >= 2018) %>%
    count(Year, data_status) %>%
    ggplot(aes(Year, n, fill = data_status)) +
    geom_col() +
    scale_fill_manual(values = status_pal) +
    labs(x = NULL, y = "Country-months", fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-gap-tiles
#| fig-cap: "Gap months per country-year (months the historical record drops as all-zero, shown as missing for the forecast layer). Blank cells are outside a country's covered span. Faceted by region."
#| fig-width: 8.5
#| fig-height: 12
#| column: body-outset

if (!is.null(panel)) {
  panel %>%
    group_by(region, iso3, Year) %>%
    summarise(gap_months = sum(data_status == "gap"), .groups = "drop") %>%
    ggplot(aes(Year, iso3, fill = gap_months)) +
    geom_tile(colour = "grey40", linewidth = 0.15) +
    facet_grid(region ~ ., scales = "free_y", space = "free_y") +
    scale_fill_gradient(low = "white", high = "#C44E52",
                        breaks = c(0, 3, 6, 9, 12), limits = c(0, 12)) +
    scale_x_continuous(breaks = seq(2010, 2026, 2), expand = c(0, 0)) +
    labs(x = NULL, y = NULL, fill = "Gap months",
         caption = "Grey background = year outside the country's record.") +
    theme(
      axis.text.y = element_text(size = 6),
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "grey82", colour = NA),
      strip.text.y = element_text(angle = 0, hjust = 0, size = 8),
      legend.position = "top"
    )
}
```

```{r}
#| label: fig-status-tiles
#| fig-cap: "Monthly data status for every country, 2010 to present. The forecast layer trains on observed + corrected + estimated months; gaps are internal missing months. Faceted by region."
#| fig-width: 9
#| fig-height: 13
#| column: body-outset

if (!is.null(panel)) {
  panel %>%
    ggplot(aes(date, iso3, fill = data_status)) +
    geom_tile() +
    facet_grid(region ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = status_pal, drop = FALSE) +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y", expand = c(0, 0)) +
    labs(x = NULL, y = NULL, fill = NULL) +
    theme(
      axis.text.y = element_text(size = 6),
      panel.grid = element_blank(),
      strip.text.y = element_text(angle = 0, hjust = 0, size = 8),
      legend.position = "top"
    )
}
```

```{r}
#| label: tbl-country-coverage
#| tbl-cap: "Per-country coverage, searchable reference (full detail in summary_country.csv)."

if (is.null(summary_country)) {
  pending("Scripts/forecasting/02_data_summary.R")
} else {
  cov_tbl <- summary_country %>%
    transmute(
      iso3, country, region,
      complete_seasons = n_complete_seasons,
      months = n_months,
      span_yrs = round(span_years, 1),
      gap_months = n_gap_months,
      pct_estimated = round(100 * pct_estimated, 1)
    ) %>%
    arrange(region, desc(complete_seasons))

  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(
      cov_tbl,
      rownames = FALSE,
      filter = "top",
      options = list(pageLength = 15, order = list(list(2, "asc")))
    )
  } else {
    knitr::kable(cov_tbl)
  }
}
```

::: {.callout-warning title="Coverage flags — countries that will constrain the backtest"}
```{r}
#| label: coverage-flags
#| output: asis

if (is.null(panel) || is.null(summary_country)) {
  pending("Scripts/forecasting/01_prepare_training_data.R and 02_data_summary.R")
} else {
  fmt_iso <- function(x) {
    if (length(x) == 0) return("_none_")
    paste0("`", paste(sort(x), collapse = "`, `"), "`")
  }

  # 1. No observed / corrected / estimated month in 2026.
  in_2026 <- panel %>%
    filter(Year == 2026, data_status != "gap") %>%
    distinct(iso3) %>%
    pull(iso3)
  no_2026 <- setdiff(unique(panel$iso3), in_2026)

  # 2. Fewer than 60 months (5 years) of actual data, whatever the span.
  thin <- summary_country %>%
    filter(n_months < 60) %>%
    arrange(n_months)

  # 3. Gap months inside the last 5 years.
  cutoff <- max(panel$date) %m-% years(5)
  gaps_5y <- panel %>%
    filter(date >= cutoff, data_status == "gap") %>%
    count(iso3, name = "gap_months") %>%
    arrange(desc(gap_months))
  heavy <- gaps_5y %>% filter(gap_months >= 12)
  light <- gaps_5y %>% filter(gap_months < 12)

  cat(sprintf(
    "**No data in 2026** (%d) -- seasonal profile only, no current-season observation; carried for history but not available for prospective scoring: %s\n\n",
    length(no_2026), fmt_iso(no_2026)
  ))

  cat(sprintf(
    "**Under 5 years of actual data** (%d) -- too thin for a fixed 5-year window; candidates to default to the seasonal baseline:\n\n",
    nrow(thin)
  ))
  cat("| iso3 | months | complete seasons | span (yr) |\n")
  cat("|------|-------:|-----------------:|----------:|\n")
  cat(paste(sprintf(
    "| `%s` | %d | %d | %.1f |",
    thin$iso3, thin$n_months, thin$n_complete_seasons, thin$span_years
  ), collapse = "\n"))
  cat("\n\n")

  cat(sprintf(
    "**Gaps in the last 5 years** (%d countries, since %s) -- a fixed-window backtest trains on discontinuous history here:\n\n",
    nrow(gaps_5y), format(cutoff)
  ))
  cat(sprintf(
    "- 12+ gap months (%d): %s\n\n",
    nrow(heavy),
    paste(sprintf("`%s` (%d)", heavy$iso3, heavy$gap_months), collapse = ", ")
  ))
  cat(sprintf("- fewer than 12 (%d): %s\n", nrow(light), fmt_iso(light$iso3)))
}
```
:::

::: {.callout-note title="Notes / observations"}
*Add: reading of the flags above — which of these default to the seasonal baseline vs still get a bespoke model, whether the expanding window rescues the gappy countries, how much of the recent record is corrected vs estimated.*
:::

------------------------------------------------------------------------

# Covariate & climate data {#sec-covariates}

Placeholder for climate and other covariate data once sourced (plan §1, item 4 — GLM+climate / GAM+climate). Candidate variables, sources, coverage by country/region, and any exploratory relationship with case counts will go here once that sourcing work starts.

::: {.callout-note appearance="simple"}
Not sourced yet — no covariate data exists in the repo at this point.
:::

::: {.callout-note title="Notes / observations"}
*Add: candidate covariates under consideration, sourcing plan, coverage.*
:::

------------------------------------------------------------------------

# Data characterisation & scenario strata {#sec-strata}

`02_data_summary.R` characterises each country's burden, seasonal shape and land area, and assigns the **provisional** scenario strata the Stage 1 hindcast will be sliced on (the remaining axis, lead time, is intrinsic to the forecast).

Seasonal shape is read from the mean seasonal profile (`DENV_average_season.csv`): `seasonal_concentration` is the Herfindahl index of the mean monthly proportions — `r round(1/12, 3)` for a flat year, rising toward 1 as a single month dominates. Land area is computed from the country outline geometry (`base_map.gpkg`), equal-area projected.

::: {.callout-important appearance="simple"}
**Provisional cut-points — to confirm with the team.**

- **seasonality_signal**: tertiles of `seasonal_concentration` across the `r if (!is.null(summary_country)) nrow(summary_country) else 96` countries → weak / moderate / strong.
- **endemicity**: *endemic* only if mean monthly cases ≥ `r endemic_min_mean_cases`, complete seasons ≥ `r endemic_min_seasons`, and zero-month fraction \< `r endemic_max_zero_fraction`; otherwise *emerging*.
- **size_class**: land area `cut()` on fixed km² breaks — `r if (exists("size_labels")) paste(size_labels, collapse = " / ") else "very_small / small / medium / large / continental"`. Change the breaks or the class count in `00_config.R`.

Alternative under consideration for seasonality: adopt Kishen Joshi's seasonality clusters in place of the tertile split (ask him for the file).
:::

```{r}
#| label: strata-crosstab-helper

# A 2-way count with a Total column and a Total row, for the small strata
# crosstabs below. `row_var` stays a factor so the row order is kept.
strata_crosstab <- function(df, row_var, col_var) {
  wide <- df %>%
    count({{ row_var }}, {{ col_var }}, .drop = FALSE) %>%
    pivot_wider(names_from = {{ col_var }}, values_from = n, values_fill = 0L) %>%
    arrange({{ row_var }}) %>%
    mutate(Total = rowSums(across(where(is.numeric))))
  total_row <- wide %>%
    summarise(across(where(is.numeric), sum)) %>%
    mutate("{{ row_var }}" := "Total", .before = 1)
  bind_rows(mutate(wide, "{{ row_var }}" := as.character({{ row_var }})), total_row)
}
```

```{r}
#| label: fig-region-seasonality
#| fig-cap: "Countries per region, split by seasonality signal (weak -> strong)."
#| fig-height: 3.6

if (!is.null(summary_country)) {
  summary_country %>%
    mutate(seasonality_signal = factor(seasonality_signal,
      levels = c("weak", "moderate", "strong"))) %>%
    count(region, seasonality_signal, .drop = FALSE) %>%
    ggplot(aes(n, reorder(region, n, sum), fill = seasonality_signal)) +
    geom_col() +
    scale_fill_brewer(palette = "YlOrRd", direction = 1) +
    labs(x = "Countries", y = NULL, fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-region-endemicity
#| fig-cap: "Countries per region, split by provisional endemicity."
#| fig-height: 3.6

if (!is.null(summary_country)) {
  summary_country %>%
    count(region, endemicity) %>%
    ggplot(aes(n, reorder(region, n, sum), fill = endemicity)) +
    geom_col() +
    scale_fill_manual(values = c(endemic = "#C44E52", emerging = "#4C72B0")) +
    labs(x = "Countries", y = NULL, fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-region-size
#| fig-cap: "Countries per region, split by land-area class (very_small -> continental)."
#| fig-height: 3.6

if (!is.null(summary_country)) {
  summary_country %>%
    mutate(size_class = factor(size_class, levels = size_labels)) %>%
    count(region, size_class, .drop = FALSE) %>%
    ggplot(aes(n, reorder(region, n, sum), fill = size_class)) +
    geom_col() +
    scale_fill_viridis_d(option = "mako", direction = -1, end = 0.92) +
    labs(x = "Countries", y = NULL, fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: tbl-strata-crosstab
#| tbl-cap: "Countries by provisional stratum: seasonality signal x endemicity."

if (is.null(summary_country)) {
  pending("Scripts/forecasting/02_data_summary.R")
} else {
  summary_country %>%
    mutate(seasonality_signal = factor(seasonality_signal,
      levels = c("weak", "moderate", "strong"))) %>%
    strata_crosstab(seasonality_signal, endemicity) %>%
    knitr::kable()
}
```

```{r}
#| label: tbl-strata-size
#| tbl-cap: "Countries by land-area class (x endemicity). Classes and km2 breaks are set in 00_config.R."

if (!is.null(summary_country)) {
  summary_country %>%
    mutate(size_class = factor(size_class, levels = size_labels)) %>%
    strata_crosstab(size_class, endemicity) %>%
    knitr::kable()
}
```

::: {.callout-note title="Notes / observations"}
**Endemicity**

- *I am surprised that the USA seems to be classified as endemic. The monthly case count min does not account for population, that maybe a problem, should it be a monthly average incidence to account for population size?*

**Seasonality**

- *seasonality tertile will be replaced Kishen's clusters as soon as I have the file*

**Size Class**

- *`size_class` is very_small-heavy (island states) — is the current 5-class / km² cut the right granularity, or collapse to small/medium/large or small/large?*

- *There are region patterns to the distribution on nation size. Will the cause confounding effects to consider?*
:::

```{r}
#| label: tbl-strata-lookup
#| tbl-cap: "Every country's assignment on all three provisional strata, with the underlying values. Searchable; full detail in strata_country.csv."

if (!is.null(strata_country)) {
  lk_tbl <- strata_country %>%
    transmute(
      iso3, country, region,
      seasonality_signal, endemicity, size_class,
      seas_conc      = round(seasonal_concentration, 3),
      mean_cases_mo  = round(mean_monthly_cases),
      seasons        = n_complete_seasons,
      area_km2       = round(country_area_km2)
    ) %>%
    arrange(region, country)

  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(
      lk_tbl,
      rownames = FALSE,
      filter = "top",
      options = list(pageLength = 15, order = list(list(2, "asc")))
    )
  } else {
    knitr::kable(lk_tbl)
  }
}
```

------------------------------------------------------------------------

# Backtest window coverage {#sec-coverage}

`03_check_window_coverage.R` is the pre-Stage-1 gate: for each monthly rolling origin and each candidate training window (fixed 5 / 7 / 10 yr, expanding), how many countries have at least `r min_train_months` valued months in the window **and** at least one target month with truth. No models are fitted.

```{r}
#| label: load-coverage

coverage_country <- read_fc("cv_splits", "rolling_origin_country.csv")

win_lvls <- c("fixed_5y", "fixed_7y", "fixed_10y", "expanding")
if (!is.null(coverage_grid)) {
  coverage_grid <- coverage_grid %>%
    mutate(
      origin_date = as.Date(origin_date),
      window_type = factor(window_type, levels = win_lvls)
    )
}
```

```{r}
#| label: fig-coverage-lines
#| fig-cap: "Countries eligible for a fit at each rolling origin, by training window. Shorter windows lose the countries with gappy recent history; all windows fall away at the final origins as the countries with no 2026 data drop out."
#| fig-height: 3.8

if (is.null(coverage_grid)) {
  pending("Scripts/forecasting/03_check_window_coverage.R")
} else {
  ggplot(coverage_grid, aes(origin_date, n_countries_eligible, colour = window_type)) +
    geom_line(linewidth = 0.7) +
    geom_hline(yintercept = unique(coverage_grid$n_countries_total),
               linetype = "dotted", colour = "grey50") +
    scale_colour_brewer(palette = "Set2") +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    labs(x = NULL, y = "Countries eligible", colour = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: tbl-coverage-summary
#| tbl-cap: "Eligible countries per window, and the first origin reaching 60."

if (!is.null(coverage_grid)) {
  coverage_grid %>%
    arrange(window_type, origin_date) %>%
    group_by(window_type) %>%
    summarise(
      first_origin  = first(n_countries_eligible),
      median_origin = round(median(n_countries_eligible)),
      last_origin   = last(n_countries_eligible),
      reaches_60    = {
        hit <- origin_date[n_countries_eligible >= 60]
        if (length(hit) == 0) "never" else format(min(hit))
      },
      .groups = "drop"
    ) %>%
    knitr::kable()
}
```

```{r}
#| label: coverage-never-eligible
#| output: asis

if (!is.null(coverage_country)) {
  cc <- coverage_country %>%
    mutate(window_type = factor(window_type, levels = win_lvls))
  all_iso <- sort(unique(cc$iso3))

  cat("**Countries never eligible at any origin** (short or gap-dominated records):\n\n")
  for (w in win_lvls) {
    ever <- cc %>% filter(window_type == w, eligible) %>% distinct(iso3) %>% pull(iso3)
    never <- setdiff(all_iso, ever)
    cat(sprintf("- `%s` (%d): %s\n", w, length(never),
                if (length(never)) paste0("`", paste(sort(never), collapse = "`, `"), "`") else "_none_"))
  }
}
```

::: {.callout-tip title="Window viability, by region"}
Median number of origins (of `r if (!is.null(coverage_grid)) length(unique(coverage_grid$origin_date)) else 121`) at which the typical country in a region is eligible, per window. `121` = every origin.

```{r}
#| label: tbl-window-viability-region

if (is.null(coverage_country)) {
  pending("Scripts/forecasting/03_check_window_coverage.R")
} else {
  guidance <- tibble::tribble(
    ~region,                              ~guidance,
    "North & Central America",            "any window; modelling choice only",
    "South America",                      "any window (only CHL needs expanding)",
    "South Asia",                         "any window",
    "East & Southeast Asia",              "any window; recent drop = no-2026-data countries, not window length",
    "Pacific Islands",                    "7 yr minimum; 10 yr / expanding covers all 15",
    "Caribbean",                          "7 yr minimum; 5 yr drops ~8 islands at recent origins",
    "Sub-Saharan Africa",                 "expanding only; short windows unusable, several still fall to Baseline 2",
    "Europe, Middle East & North Africa", "no standalone stratum under any window; pool to global"
  )

  n_ctry <- coverage_country %>% distinct(region, iso3) %>% count(region, name = "n")

  coverage_country %>%
    mutate(window_type = factor(window_type, levels = win_lvls)) %>%
    group_by(region, window_type, iso3) %>%
    summarise(n_elig = sum(eligible), .groups = "drop") %>%
    group_by(region, window_type) %>%
    summarise(median_origins = round(median(n_elig)), .groups = "drop") %>%
    pivot_wider(names_from = window_type, values_from = median_origins) %>%
    left_join(n_ctry, by = "region") %>%
    left_join(guidance, by = "region") %>%
    mutate(region = factor(region, levels = region_levels)) %>%
    arrange(region) %>%
    transmute(
      region = paste0(as.character(region), " (", n, ")"),
      `5 yr` = fixed_5y, `7 yr` = fixed_7y, `10 yr` = fixed_10y, expanding,
      guidance
    ) %>%
    knitr::kable(align = c("l", "r", "r", "r", "r", "l"))
}
```

**Default: expanding** — the only window viable everywhere it physically can be (94 / 96 countries eligible at some origin). **`fixed_10y`** carried as the trimming comparison for the data-rich regions (≈ expanding for coverage, diverges only where \>10 yr of history exists). **`fixed_7y`** as a Caribbean / Pacific fallback. **`fixed_5y`** diagnostic only — it cannot be a global default.
:::

::: {.callout-note title="Notes / observations"}
*Add: final window decision, how the chronically-intermittent countries (SSA, Caribbean tail) are handled, whether fixed-5yr is worth the compute as a diagnostic.*
:::

------------------------------------------------------------------------

# Model progress {#sec-progress}

Which models have a definition file, and which stages they've cleared so far.

```{r}
#| label: tbl-model-progress
#| tbl-cap: "Stage 0 status for every model with a Stage 0 record."

if (is.null(stage0_diag)) {
  pending("Scripts/forecasting/run_stage0_fit.R")
} else {
  scored_models <- if (!is.null(stage1_scores_by_lead)) unique(stage1_scores_by_lead$model) else character(0)

  stage0_diag %>%
    transmute(
      Model = model,
      `Run at` = run_at,
      `Stage 0` = if_else(pass, "PASS", "FAIL"),
      `Stage 1` = if_else(model %in% scored_models, "scored", "-"),
      `Stage 2` = "-",
      Notes = notes
    ) %>%
    knitr::kable()
}
```

```{r}
#| label: model-files-note
#| output: asis

model_files <- sort(list.files(
  file.path(repo_root, "Scripts", "forecasting", "models"),
  pattern = "^[0-9].*\\.R$"
))
have_record <- if (!is.null(stage0_diag)) nrow(stage0_diag) else 0L
cat(sprintf(
  "%d model definition file%s in `Scripts/forecasting/models/`: %s. %d have a Stage 0 record above; the rest are defined but not yet run (or parked).\n",
  length(model_files), if (length(model_files) == 1) "" else "s",
  paste0("`", model_files, "`", collapse = ", "),
  have_record
))
```

::: {.callout-note title="Notes / observations"}
*Add: which parked/unrun models need attention before Stage 1.*
:::

------------------------------------------------------------------------

# Stage 0 — model fit diagnostics {#sec-stage0}

`run_stage0_fit.R` fits each model once on the full panel and runs its diagnostic checklist (convergence, dispersion, residual structure). Diagnostic only — no skill claim.

```{r}
#| label: tbl-stage0
#| tbl-cap: "Stage 0 fit diagnostics."

if (is.null(stage0_diag)) {
  pending("Scripts/forecasting/run_stage0_fit.R")
} else {
  knitr::kable(stage0_diag)
}
```

::: {.callout-note title="Notes / observations"}
*Add: did every candidate pass the sanity check? Any coefficient signs or dispersion values worth flagging before Stage 1.*
:::

------------------------------------------------------------------------

# Stage 1 — hindcast skill {#sec-stage1}

Rolling-origin backtest (`run_stage1_hindcast.R`), scored by `run_stage1_score.R`. Reads top to bottom, widest view first: model performance overall, then by lead time, then by scenario category, then lead x category together; calibration and the full data table sit at the end for reference.

| Role | Metric | Note |
|----|----|----|
| Primary | CRPS (log scale) | `wis`/`wis_log` dropped 14-09-2026, confirmed identical to `crps`/`crps_log` at today's 5-quantile schema. `scoringutils`' own pairwise-comparison output still calls it `wis` internally. |
| Secondary | uMAE, uRMSE | GDO's burden-normalised convention |
| Secondary | PIT / interval coverage | 50%, 90% |
| Operational | Peak-timing difference | Months; D-MOSS (Campbell et al. 2026) convention |
| Operational | DTW distance | Path-length-normalised (`normalizedDistance`); `NA` below 3 resolved months |

`n_seasons` (distinct `iso3 x season` pairs, not raw row count) and `thin_evidence = n_seasons < 5` are carried on every table/figure below — flagged, never dropped.

```{r}
#| label: sec-stage1-guard

stage1_ready <- !is.null(stage1_scores_by_lead)
if (!stage1_ready) pending("Scripts/forecasting/run_stage1_hindcast.R and run_stage1_score.R")

stratum_axes <- c("seasonality_signal", "endemicity", "size_class")
```

```{r}
#| label: tile-heatmap-helper

# One heatmap style, reused for CRPS-by-category-x-lead below and for any
# later ad hoc / cross-strata comparison (matches fig-nowcast-uncertainty-region's
# look). Thin evidence is a black tile border, not an alpha fade - fading
# competed with the fill's own colour intensity and was hard to read.
tile_heatmap <- function(df, x_var, y_var, fill_var, x_lab, fill_lab,
                         facet_var = "model", thin_var = NULL) {
  p <- ggplot(df, aes(.data[[x_var]], .data[[y_var]], fill = .data[[fill_var]]))
  p <- if (is.null(thin_var)) {
    p + geom_tile(colour = "white")
  } else {
    p + geom_tile(aes(colour = .data[[thin_var]]), linewidth = 0.6) +
      scale_colour_manual(values = c(`TRUE` = "black", `FALSE` = "white"), guide = "none")
  }
  p + facet_wrap(stats::reformulate(facet_var)) +
    scale_fill_viridis_c(option = "magma", direction = -1, na.value = "grey90") +
    labs(x = x_lab, y = NULL, fill = fill_lab)
}
```

### Overview: model x window

Everything pooled — the widest possible view.

```{r}
#| label: fig-overview
#| fig-cap: "Primary and operational scores, pooled across lead time and scenario category."
#| fig-height: 3.2

if (stage1_ready) {
  bind_rows(
    stage1_scores_row %>%
      group_by(model, window_type) %>%
      summarise(value = mean(crps_log, na.rm = TRUE), .groups = "drop") %>%
      mutate(metric = "CRPS (log)"),
    stage1_operational %>%
      filter(stratum_axis == "overall") %>%
      transmute(model, window_type, value = mean_peak_timing_diff, metric = "Peak-timing (mo)"),
    stage1_operational %>%
      filter(stratum_axis == "overall") %>%
      transmute(model, window_type, value = mean_dtw_distance, metric = "DTW (norm.)")
  ) %>%
    mutate(metric = factor(metric, levels = c("CRPS (log)", "Peak-timing (mo)", "DTW (norm.)"))) %>%
    ggplot(aes(model, value, fill = window_type)) +
    geom_col(position = "dodge") +
    facet_wrap(~metric, scales = "free_y") +
    scale_fill_brewer(palette = "Set2") +
    labs(x = NULL, y = NULL, fill = "Window") +
    theme(legend.position = "top")
}
```

### By lead time

```{r}
#| label: fig-crps-by-lead
#| fig-cap: "CRPS (log scale) by lead time and window."
#| fig-height: 3.6

if (stage1_ready) {
  stage1_scores_by_lead %>%
    ggplot(aes(lead_time, crps_log, colour = model, group = model)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.8) +
    facet_wrap(~window_type) +
    scale_colour_brewer(palette = "Set2") +
    scale_x_continuous(breaks = sort(unique(stage1_scores_by_lead$lead_time))) +
    labs(x = "Lead time (months)", y = "CRPS (log scale)", colour = NULL) +
    theme(legend.position = "top")
}
```

Same comparison as a ratio to `nowcast` (GDO's mandatory seasonal-profile baseline) — `< 1` beats it. `leaderboard.csv` also carries `glm_ar1` as a comparator and the full `scoringutils` pairwise output (p-values, the scaled relative-skill version) for once more models exist.

```{r}
#| label: fig-leaderboard
#| fig-cap: "CRPS relative to nowcast, by lead time and window."
#| fig-height: 3.2

if (stage1_ready) {
  # Both `baseline == "nowcast"` AND `compare_against == "nowcast"`: leaderboard.csv
  # has one row per (model, compare_against) pair, so filtering on `baseline`
  # alone leaves nowcast's own comparisons against every OTHER model in too -
  # duplicate x-values per model that broke the line into a zigzag.
  stage1_leaderboard %>%
    filter(baseline == "nowcast", compare_against == "nowcast") %>%
    ggplot(aes(horizon, mean_scores_ratio, colour = model, group = model)) +
    geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.8) +
    facet_wrap(~window_type) +
    scale_colour_brewer(palette = "Set2") +
    scale_x_continuous(breaks = sort(unique(stage1_leaderboard$horizon))) +
    labs(x = "Lead time (months)", y = "CRPS relative to nowcast", colour = NULL) +
    theme(legend.position = "top")
}
```

Peak-timing and DTW have no lead-time axis (both are computed once per full 1-6 month trajectory, not per lead — see the metrics table above); the closest analogue is the calendar month the forecast was *issued* from, following Campbell et al.'s own month-based view.

```{r}
#| label: fig-operational-by-month
#| fig-cap: "Peak-timing difference and DTW distance by origin month, per model and window (Campbell et al. 2026 style)."
#| fig-height: 4.2

if (!is.null(stage1_operational)) {
  stage1_operational_row <- read_fc("stage1_hindcast", "scores", "operational_row.csv")
}
if (!is.null(stage1_operational_row)) {
  stage1_operational_row %>%
    mutate(origin_month = lubridate::month(as.Date(origin_date))) %>%
    group_by(model, window_type, origin_month) %>%
    summarise(
      `Peak-timing diff (mo)` = mean(peak_timing_diff, na.rm = TRUE),
      `DTW (normalised)`      = mean(dtw_distance, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    pivot_longer(c(`Peak-timing diff (mo)`, `DTW (normalised)`), names_to = "metric", values_to = "value") %>%
    ggplot(aes(origin_month, value, colour = model, group = model)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.6) +
    facet_grid(metric ~ window_type, scales = "free_y") +
    scale_colour_brewer(palette = "Set2") +
    scale_x_continuous(breaks = 1:12) +
    labs(x = "Origin month", y = NULL, colour = NULL) +
    theme(legend.position = "top")
} else {
  pending("Scripts/forecasting/run_stage1_score.R")
}
```

### By scenario category

Pooled across lead time, one axis at a time.

```{r}
#| label: fig-crps-by-category
#| fig-cap: "CRPS (log scale) by scenario category, pooled across lead time."
#| fig-height: 3.4

if (stage1_ready) {
  stage1_scores_by_stratum %>%
    group_by(model, window_type, stratum_axis, stratum_value) %>%
    summarise(crps_log = weighted.mean(crps_log, w = n_obs, na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(stratum_value, crps_log, fill = model)) +
    geom_col(position = "dodge") +
    facet_wrap(~stratum_axis, scales = "free_x") +
    scale_fill_brewer(palette = "Set2") +
    labs(x = NULL, y = "CRPS (log scale)", fill = NULL) +
    theme(legend.position = "top", axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r}
#| label: fig-operational-by-category
#| fig-cap: "Peak-timing difference and DTW distance by scenario category, pooled across lead time."
#| fig-height: 3.6

if (!is.null(stage1_operational)) {
  stage1_operational %>%
    filter(stratum_axis != "overall") %>%
    select(model, window_type, stratum_axis, stratum_value,
           `Peak-timing diff (mo)` = mean_peak_timing_diff, `DTW (normalised)` = mean_dtw_distance) %>%
    pivot_longer(c(`Peak-timing diff (mo)`, `DTW (normalised)`), names_to = "metric", values_to = "value") %>%
    ggplot(aes(stratum_value, value, fill = model)) +
    geom_col(position = "dodge") +
    facet_grid(metric ~ stratum_axis, scales = "free", space = "free_x") +
    scale_fill_brewer(palette = "Set2") +
    labs(x = NULL, y = NULL, fill = NULL) +
    theme(legend.position = "top", axis.text.x = element_text(angle = 20, hjust = 1))
} else {
  pending("Scripts/forecasting/run_stage1_score.R")
}
```

### By category x lead time

Same view as `fig-nowcast-uncertainty-region`, generalised: category on the y-axis, lead time on the x-axis, faceted by model. Reusable (`tile_heatmap()` above) for any later ad hoc or cross-strata comparison.

```{r}
#| label: fig-score-heatmap
#| fig-cap: "CRPS (log scale) by scenario category and lead time, faceted by model. Black border = thin evidence (n_seasons < 5)."
#| fig-height: 6

if (stage1_ready) {
  tile_heatmap(
    stage1_scores_by_stratum, x_var = "lead_time", y_var = "stratum_value",
    fill_var = "crps_log", x_lab = "Lead time (months)", fill_lab = "CRPS\n(log)",
    thin_var = "thin_evidence"
  ) +
    facet_grid(stratum_axis ~ model, scales = "free_y", space = "free_y")
}
```

```{r}
#| label: fig-operational-heatmap
#| fig-cap: "Peak-timing difference by scenario category and origin month, faceted by model."
#| fig-height: 6

if (!is.null(stage1_operational_row)) {
  op_cat <- bind_rows(lapply(stratum_axes, function(axis) {
    stage1_operational_row %>%
      rename(stratum_value = all_of(axis)) %>%
      mutate(origin_month = lubridate::month(as.Date(origin_date))) %>%
      group_by(model, window_type, stratum_value, origin_month) %>%
      summarise(
        n_seasons = n_distinct(paste(iso3, season)),
        mean_peak_timing_diff = mean(peak_timing_diff, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(stratum_axis = axis, thin_evidence = n_seasons < 5)
  }))

  tile_heatmap(
    op_cat, x_var = "origin_month", y_var = "stratum_value",
    fill_var = "mean_peak_timing_diff", x_lab = "Origin month", fill_lab = "Peak-timing\ndiff (mo)",
    thin_var = "thin_evidence"
  ) +
    facet_grid(stratum_axis ~ model, scales = "free_y", space = "free_y") +
    scale_x_continuous(breaks = 1:12)
} else {
  pending("Scripts/forecasting/run_stage1_score.R")
}
```

### Calibration: PIT and coverage

```{r}
#| label: fig-pit
#| fig-cap: "PIT histogram by model and window. A well-calibrated model is flat across bins."
#| fig-height: 3.4

if (!is.null(stage1_pit)) {
  # scoringutils' PIT bins are NOT equal-width (5%/20%/25%/25%/20%/5%) - parse
  # the true edges out of `bin` (e.g. "[0.05,0.25)") rather than drawing
  # equal-width bars at `mid`, which would mis-represent the wider middle
  # bins and overlap the narrow end bins.
  pit_rect <- stage1_pit %>%
    mutate(
      xmin = as.numeric(str_extract(bin, "(?<=\\[)[0-9.]+")),
      xmax = as.numeric(str_extract(bin, "[0-9.]+(?=[\\]\\)])"))
    )
  ggplot(pit_rect, aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = density)) +
    geom_rect(fill = "#4C72B0", colour = "white") +
    geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
    facet_grid(window_type ~ model) +
    labs(x = "PIT", y = "Density")
} else {
  pending("Scripts/forecasting/run_stage1_score.R")
}
```

```{r}
#| label: fig-coverage
#| fig-cap: "Empirical vs nominal quantile coverage, by model and window. Points on the diagonal are well-calibrated; below it is over-confident (too narrow)."
#| fig-height: 3.6

if (!is.null(stage1_coverage)) {
  ggplot(stage1_coverage, aes(quantile_level, quantile_coverage, colour = model)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.6) +
    facet_wrap(~window_type) +
    scale_colour_brewer(palette = "Set2") +
    labs(x = "Nominal quantile", y = "Empirical coverage", colour = NULL) +
    theme(legend.position = "top")
} else {
  pending("Scripts/forecasting/run_stage1_score.R")
}
```

### Full stratum table (reference)

Every model x window x lead x category cell, long format, searchable — the detail behind every figure above.

```{r}
#| label: tbl-scores-by-stratum
#| tbl-cap: "Stage 1 scores by scenario stratum. Thin-evidence cells (n_seasons < 5) flagged, not dropped."

if (stage1_ready) {
  strat_tbl <- stage1_scores_by_stratum %>%
    transmute(
      Model = model, Window = window_type, Axis = stratum_axis, Stratum = stratum_value,
      Lead = lead_time, n = n_obs, n_seasons,
      thin = thin_evidence,
      CRPS = round(crps, 1), `CRPS (log)` = round(crps_log, 3),
      `Cov. 90%` = round(coverage_90, 2), uMAE = round(uMAE, 3), uRMSE = round(uRMSE, 3)
    ) %>%
    arrange(Axis, Stratum, Model, Window, Lead)

  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(
      strat_tbl, rownames = FALSE, filter = "top",
      options = list(pageLength = 15)
    ) %>%
      DT::formatStyle("thin", target = "row",
                      backgroundColor = DT::styleEqual(c(TRUE, FALSE), c("#fdf3d0", "white")))
  } else {
    knitr::kable(strat_tbl)
  }
}
```

::: {.callout-note title="Notes / observations"}
*Add as results accumulate on the full backtest grid (this is currently the reduced check run — one window, stride-6 origins). Watch in particular: `coverage_90` degrading at longer leads below the \~0.7 floor you'd want from a well-calibrated model; whether that holds up once the full grid runs.*
:::

------------------------------------------------------------------------

# Conclusions & manuscript notes {#sec-conclusions}

Running synthesis for the forecasting paper (target: PLOS Comp Biol / PLOS NTD).

- **Data.** *...*
- **Strata.** *...*
- **Baselines.** *...*
- **Best model(s) so far.** *...*
- **Where the forecast is trustworthy / where it is not.** *...*

------------------------------------------------------------------------

# Open questions & directions {#sec-directions}

- Minimum complete-seasons floor below which a country defaults to the seasonal baseline (`[Q2.4]`).
- Spatial (region-grouped) blocking alongside the temporal windows (`§6`).
- Promotion rule: beat both baselines at *majority* vs *all* lead times (`§6`).
- Refit cadence for the hierarchical model — monthly vs quarterly (`[Q2.6]`).
- Operational-utility scoring design at \~83-country scale (`[Q5.2]`).
- Climate covariate sources and the GAM / hierarchical model structure.

------------------------------------------------------------------------

# Appendix: Model details {#sec-model-details}

### Forecast: August-December 2026

Case counts, not skill — every candidate model's actual forecast from the current data edge, one plot per region, faceted by model so a region's behaviour is directly comparable across models. Solid = observed 2026 to date; dashed + band = forecast (median, 90%), from the Stage 0 fit's origin (the panel's last observed month).

```{r}
#| label: forecast-2026-load

# Stage 0 wiring-check forecast for every model - this IS the current
# production forecast (fit on the full panel, origin = its last observed
# month), not a hindcast-grid artifact. Both models share one origin, but
# that's not assumed - checked below.
model_wiring_forecasts <- function(slug) {
  fc <- read_fc("stage0_fit", slug, "wiring_check_forecasts.csv")
  if (is.null(fc)) return(NULL)
  fc %>%
    mutate(origin_date = as.Date(origin_date), target_date = as.Date(target_date)) %>%
    left_join(region_lookup, by = "iso3") %>%
    mutate(model = slug, .before = 1)
}

wiring_models <- c("nowcast", "glm_ar1")
wiring_all <- bind_rows(lapply(wiring_models, model_wiring_forecasts))
```

```{r}
#| label: forecast-2026-data

if (!is.null(wiring_all) && nrow(wiring_all) > 0) {
  origins <- unique(wiring_all$origin_date)
  #---check: every model's Stage 0 fit shares one origin - if not, the anchor
  # logic below (one shared `obs_2026`) would silently misalign per model.
  if (length(origins) > 1) {
    cli::cli_abort("Models disagree on Stage 0 origin: {paste(format(origins), collapse=', ')}.")
  }
  fc_origin <- origins

  # Window fixed to what was asked for (Aug-Dec 2026), even though the
  # model's own horizon runs one month further (to fc_origin + 6).
  window_start <- as.Date("2026-08-01")
  window_end   <- as.Date("2026-12-01")

  obs_2026 <- panel %>%
    filter(Year == 2026, !is.na(cases)) %>%
    transmute(iso3, region = as.character(region), date, value = cases,
              segment = "observed", lower90 = NA_real_, upper90 = NA_real_) %>%
    tidyr::crossing(model = wiring_models)

  # Anchor the forecast segment at the origin's observed value, per model, so
  # the dashed line starts exactly where the solid line ends.
  fc_anchor <- panel %>%
    filter(date == fc_origin, !is.na(cases)) %>%
    transmute(iso3, region = as.character(region), date, value = cases,
              segment = "forecast", lower90 = cases, upper90 = cases) %>%
    tidyr::crossing(model = wiring_models)

  fc_future <- wiring_all %>%
    filter(!is.na(.pred), target_date >= window_start, target_date <= window_end) %>%
    transmute(model, iso3, region = as.character(region), date = target_date,
              value = .pred, segment = "forecast",
              lower90 = .pred_lower90, upper90 = .pred_upper90)

  forecast_2026_df <- bind_rows(obs_2026, fc_anchor, fc_future) %>%
    mutate(region = factor(region, levels = region_levels)) %>%
    arrange(model, iso3, date)

  forecast_2026_labels <- forecast_2026_df %>%
    filter(!is.na(value)) %>%
    group_by(model, iso3, region) %>%
    slice_max(date, n = 1, with_ties = FALSE) %>%
    ungroup()
}
```

```{r}
#| label: fig-forecast-2026-by-region
#| output: asis
#| fig-cap: "August-December 2026 forecast, by region, faceted by model."

if (is.null(wiring_all) || nrow(wiring_all) == 0) {
  pending("Scripts/forecasting/run_stage0_fit.R")
} else {
  fig_dir <- tempfile("forecast_2026_region_figs_")
  dir.create(fig_dir)

  for (rg in levels(forecast_2026_df$region)) {
    d <- forecast_2026_df %>% filter(region == rg)
    if (nrow(d) == 0) next
    lbl <- forecast_2026_labels %>% filter(region == rg)
    n_ctry <- n_distinct(d$iso3)

    p <- ggplot(d, aes(date, value, group = iso3, colour = iso3)) +
      geom_ribbon(
        data = d %>% filter(segment == "forecast", !is.na(lower90)),
        aes(ymin = lower90, ymax = upper90, fill = iso3),
        colour = NA, alpha = 0.12, show.legend = FALSE
      ) +
      geom_line(data = d %>% filter(segment == "observed"), linewidth = 0.6, linetype = "solid") +
      geom_line(data = d %>% filter(segment == "forecast"), linewidth = 0.6, linetype = "22") +
      ggrepel::geom_text_repel(
        data = lbl, aes(label = iso3), size = 3, hjust = 0, direction = "y",
        nudge_x = 15, segment.size = 0.2, min.segment.length = 0,
        max.overlaps = Inf, show.legend = FALSE
      ) +
      facet_wrap(~model) +
      scale_colour_viridis_d(option = "turbo") +
      scale_fill_viridis_d(option = "turbo") +
      scale_x_date(date_labels = "%b", expand = expansion(mult = c(0.02, 0.15))) +
      scale_y_continuous(labels = scales::label_comma()) +
      labs(x = NULL, y = "Monthly cases")

    fname <- file.path(fig_dir, paste0(make.names(rg), ".png"))
    fig_h <- max(3.5, 0.25 * n_ctry + 2.2)
    ggsave(fname, p, width = 10, height = fig_h, dpi = 130)

    cat(sprintf(
      "\n\n#### %s\n\n<img src=\"%s\" style=\"width:100%%;\" alt=\"Aug-Dec 2026 forecast for %s, by model\">\n\n",
      rg, knitr::image_uri(fname), rg
    ))
  }
}
```

------------------------------------------------------------------------

# Provenance {.unnumbered}

```{r}
#| label: provenance
#| echo: false

tibble::tibble(
  item = c("Data snapshot", "Forecast horizons", "Rolling windows (months)",
           "Min training months", "Origin start", "Rendered"),
  value = c(
    snapshot_date,
    paste(range(forecast_horizons), collapse = "-"),
    paste(names(roll_windows), unname(roll_windows), sep = "=", collapse = ", "),
    as.character(min_train_months),
    format(origin_start),
    format(Sys.time(), "%Y-%m-%d %H:%M")
  )
) %>% knitr::kable()
```

```{r}
#| label: sessioninfo
#| code-fold: true
#| code-summary: "R session"

sessionInfo()
```
