---
title: "GDO Forecasting — Evaluation Notebook"
subtitle: "Results, figures and working notes for the forecasting objective"
author: "K M Susong"
date: today
format:
  html:
    grid:
      body-width: 1000px
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(lubridate)
library(ggplot2)
library(knitr)

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 4.5, dpi = 150
)

theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold")))

# --- Locate the repo root, whatever directory this renders from ---
find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/")
  repeat {
    if (file.exists(file.path(d, "Scripts", "V1_Pipeline.R"))) return(d)
    parent <- dirname(d)
    if (identical(parent, d)) {
      stop("DENV_global_observatory repo root not found from ", start)
    }
    d <- parent
  }
}
repo_root <- find_repo_root()

# Pinned parameters and provisional thresholds come straight from the pipeline
# config, so this notebook always reports what the scripts actually used.
source(file.path(repo_root, "Scripts", "forecasting", "00_config.R"), local = TRUE)

fc_out <- file.path(repo_root, forecast_out)

# Load a forecasting output if it exists yet, else NULL.
read_fc <- function(...) {
  p <- file.path(fc_out, ...)
  if (!file.exists(p)) return(NULL)
  readr::read_csv(p, show_col_types = FALSE)
}

# Emit a "not generated yet" callout for sections whose script has not run.
# asis_output renders correctly whether or not the calling chunk is `asis`.
pending <- function(script) {
  knitr::asis_output(sprintf(
    "::: {.callout-note appearance=\"simple\"}\nNot generated yet -- run `%s` and re-render.\n:::\n",
    script
  ))
}
```

```{r}
#| label: load-outputs

panel          <- read_fc("training_data", paste0("training_panel_", snapshot_date, ".csv"))
region_lookup  <- read_fc("training_data", "country_region_lookup.csv")
summary_country <- read_fc("data_summary", "summary_country.csv")
summary_region  <- read_fc("data_summary", "summary_region.csv")
strata_country  <- read_fc("data_summary", "strata_country.csv")
coverage_grid   <- read_fc("cv_splits", "rolling_origins.csv")
stage0_diag     <- read_fc("stage0_fit", "diagnostics.csv")

if (!is.null(panel)) {
  panel <- panel %>%
    mutate(
      date = as.Date(date),
      data_status = factor(data_status,
        levels = c("observed", "corrected", "estimated", "gap"))
    )
}
```

## About this notebook {.unnumbered}

This is the **results record** for the GDO forecasting objective — the place to
look at what each pipeline step produced, keep the figures and tables that will
go into the manuscript, and write down what they mean and where the work goes
next.

It is deliberately **not**:

- the running task log — that lives in Logseq;
- a home for pipeline code — the training-data build, data summary, model
  definitions, runners and scoring pipeline are all under
  `Scripts/forecasting/`. This notebook only *reads their outputs*.

::: {.callout-note appearance="simple"}
**Pinned data snapshot:** `Output/`r snapshot_date`/` — merged historical record
plus current-period nowcast output. Every result below is against this frozen
snapshot; re-pin in `Scripts/forecasting/00_config.R` and re-run the pipeline to
refresh.
:::

Plan of record: `planning/2026_forecast_evaluation/03_Evaluation_Plan_DRAFT`.
Stages: **0** in-sample fit diagnostics · **1** rolling-origin hindcast
(5 / 7 / 10 yr + expanding) · **2** prospective shadow deployment.

---

# Training data {#sec-training-data}

`01_prepare_training_data.R` merges the deep historical record
(`full_data_season_monthly_proportions.csv`, complete seasons) with the
current-period nowcast output (`DENV_cases_nowcast_output.csv`). The nowcast view
wins on any overlapping country-month, so the panel runs as far forward as data
exists and carries the post-nowcast "ground truth".

`data_status` is preserved throughout: **observed** (raw report), **corrected**
(reporting-delay factor applied), **estimated** (seasonal-proportion nowcast),
**gap** (internal month with no value — the historical file drops all-zero
years).

```{r}
#| label: status-palette

status_pal <- c(
  observed  = "#4C72B0",
  corrected = "#DD8452",
  estimated = "#C44E52",
  gap       = "grey85"
)

if (!is.null(panel)) {
  panel <- panel %>%
    mutate(region = factor(region, levels = region_levels))
}
```

```{r}
#| label: tbl-panel-overview
#| tbl-cap: "Training panel: coverage and composition."

if (is.null(panel)) {
  pending("Scripts/forecasting/01_prepare_training_data.R")
} else {
  fmt <- function(x) formatC(x, big.mark = ",", format = "d")
  tibble::tibble(
    Metric = c("Countries", "Country-months", "Date range",
               "Observed", "Corrected (reporting-delay factor)",
               "Estimated (seasonal-proportion nowcast)",
               "Gap (internal month, no value)"),
    Value = c(
      fmt(n_distinct(panel$iso3)),
      fmt(nrow(panel)),
      paste(format(min(panel$date)), "to", format(max(panel$date))),
      fmt(sum(panel$data_status == "observed")),
      fmt(sum(panel$data_status == "corrected")),
      fmt(sum(panel$data_status == "estimated")),
      fmt(sum(panel$data_status == "gap"))
    )
  ) %>% knitr::kable(align = c("l", "r"))
}
```

```{r}
#| label: fig-data-status-year
#| fig-cap: "Data-status composition by calendar year. Recent years carry the reporting-delay corrections and seasonal-proportion estimates the forecast layer treats as ground truth."
#| fig-height: 3.6

if (!is.null(panel)) {
  panel %>%
    filter(Year >= 2018) %>%
    count(Year, data_status) %>%
    ggplot(aes(Year, n, fill = data_status)) +
    geom_col() +
    scale_fill_manual(values = status_pal) +
    labs(x = NULL, y = "Country-months", fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-gap-tiles
#| fig-cap: "Gap months per country-year (months the historical record drops as all-zero, shown as missing for the forecast layer). Blank cells are outside a country's covered span. Faceted by region."
#| fig-width: 8.5
#| fig-height: 12
#| column: body-outset

if (!is.null(panel)) {
  panel %>%
    group_by(region, iso3, Year) %>%
    summarise(gap_months = sum(data_status == "gap"), .groups = "drop") %>%
    ggplot(aes(Year, iso3, fill = gap_months)) +
    geom_tile(colour = "grey40", linewidth = 0.15) +
    facet_grid(region ~ ., scales = "free_y", space = "free_y") +
    scale_fill_gradient(low = "white", high = "#C44E52",
                        breaks = c(0, 3, 6, 9, 12), limits = c(0, 12)) +
    scale_x_continuous(breaks = seq(2010, 2026, 2), expand = c(0, 0)) +
    labs(x = NULL, y = NULL, fill = "Gap months",
         caption = "Grey background = year outside the country's record.") +
    theme(
      axis.text.y = element_text(size = 6),
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "grey82", colour = NA),
      strip.text.y = element_text(angle = 0, hjust = 0, size = 8),
      legend.position = "top"
    )
}
```

```{r}
#| label: fig-status-tiles
#| fig-cap: "Monthly data status for every country, 2010 to present. The forecast layer trains on observed + corrected + estimated months; gaps are internal missing months. Faceted by region."
#| fig-width: 9
#| fig-height: 13
#| column: body-outset

if (!is.null(panel)) {
  panel %>%
    ggplot(aes(date, iso3, fill = data_status)) +
    geom_tile() +
    facet_grid(region ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = status_pal, drop = FALSE) +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y", expand = c(0, 0)) +
    labs(x = NULL, y = NULL, fill = NULL) +
    theme(
      axis.text.y = element_text(size = 6),
      panel.grid = element_blank(),
      strip.text.y = element_text(angle = 0, hjust = 0, size = 8),
      legend.position = "top"
    )
}
```

```{r}
#| label: tbl-country-coverage
#| tbl-cap: "Per-country coverage, searchable reference (full detail in summary_country.csv)."

if (is.null(summary_country)) {
  pending("Scripts/forecasting/02_data_summary.R")
} else {
  cov_tbl <- summary_country %>%
    transmute(
      iso3, country, region,
      complete_seasons = n_complete_seasons,
      months = n_months,
      span_yrs = round(span_years, 1),
      gap_months = n_gap_months,
      pct_estimated = round(100 * pct_estimated, 1)
    ) %>%
    arrange(region, desc(complete_seasons))

  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(
      cov_tbl,
      rownames = FALSE,
      filter = "top",
      options = list(pageLength = 15, order = list(list(2, "asc")))
    )
  } else {
    knitr::kable(cov_tbl)
  }
}
```

::: {.callout-warning title="Coverage flags — countries that will constrain the backtest"}

```{r}
#| label: coverage-flags
#| output: asis

if (is.null(panel) || is.null(summary_country)) {
  pending("Scripts/forecasting/01_prepare_training_data.R and 02_data_summary.R")
} else {
  fmt_iso <- function(x) {
    if (length(x) == 0) return("_none_")
    paste0("`", paste(sort(x), collapse = "`, `"), "`")
  }

  # 1. No observed / corrected / estimated month in 2026.
  in_2026 <- panel %>%
    filter(Year == 2026, data_status != "gap") %>%
    distinct(iso3) %>%
    pull(iso3)
  no_2026 <- setdiff(unique(panel$iso3), in_2026)

  # 2. Fewer than 60 months (5 years) of actual data, whatever the span.
  thin <- summary_country %>%
    filter(n_months < 60) %>%
    arrange(n_months)

  # 3. Gap months inside the last 5 years.
  cutoff <- max(panel$date) %m-% years(5)
  gaps_5y <- panel %>%
    filter(date >= cutoff, data_status == "gap") %>%
    count(iso3, name = "gap_months") %>%
    arrange(desc(gap_months))
  heavy <- gaps_5y %>% filter(gap_months >= 12)
  light <- gaps_5y %>% filter(gap_months < 12)

  cat(sprintf(
    "**No data in 2026** (%d) -- seasonal profile only, no current-season observation; carried for history but not available for prospective scoring: %s\n\n",
    length(no_2026), fmt_iso(no_2026)
  ))

  cat(sprintf(
    "**Under 5 years of actual data** (%d) -- too thin for a fixed 5-year window; candidates to default to the seasonal baseline:\n\n",
    nrow(thin)
  ))
  cat("| iso3 | months | complete seasons | span (yr) |\n")
  cat("|------|-------:|-----------------:|----------:|\n")
  cat(paste(sprintf(
    "| `%s` | %d | %d | %.1f |",
    thin$iso3, thin$n_months, thin$n_complete_seasons, thin$span_years
  ), collapse = "\n"))
  cat("\n\n")

  cat(sprintf(
    "**Gaps in the last 5 years** (%d countries, since %s) -- a fixed-window backtest trains on discontinuous history here:\n\n",
    nrow(gaps_5y), format(cutoff)
  ))
  cat(sprintf(
    "- 12+ gap months (%d): %s\n\n",
    nrow(heavy),
    paste(sprintf("`%s` (%d)", heavy$iso3, heavy$gap_months), collapse = ", ")
  ))
  cat(sprintf("- fewer than 12 (%d): %s\n", nrow(light), fmt_iso(light$iso3)))
}
```

:::

::: {.callout-note title="Notes / observations"}
_Add: reading of the flags above — which of these default to the seasonal
baseline vs still get a bespoke model, whether the expanding window rescues the
gappy countries, how much of the recent record is corrected vs estimated._
:::

---

# Data characterisation & scenario strata {#sec-strata}

`02_data_summary.R` characterises each country's burden and seasonal shape, and
assigns the two **provisional** scenario strata the Stage 1 backtest will be
sliced on (the third axis, lead time, is intrinsic to the forecast).

Seasonal shape is read from the mean seasonal profile (`DENV_average_season.csv`):
`seasonal_concentration` is the Herfindahl index of the mean monthly proportions
— `r round(1/12, 3)` for a flat year, rising toward 1 as a single month
dominates.

::: {.callout-important appearance="simple"}
**Provisional cut-points — to confirm with the team.**

- **seasonality_signal**: tertiles of `seasonal_concentration` across the
  `r if (!is.null(summary_country)) nrow(summary_country) else 96` countries →
  weak / moderate / strong.
- **endemicity**: *endemic* only if mean monthly cases ≥
  `r endemic_min_mean_cases`, complete seasons ≥ `r endemic_min_seasons`, and
  zero-month fraction < `r endemic_max_zero_fraction`; otherwise *emerging*.

Alternative under consideration: adopt the seasonality clusters from Joshi et al.
in place of the tertile split.
:::

```{r}
#| label: tbl-strata-crosstab
#| tbl-cap: "Countries by provisional stratum."

if (is.null(summary_country)) {
  pending("Scripts/forecasting/02_data_summary.R")
} else {
  summary_country %>%
    count(seasonality_signal, endemicity) %>%
    pivot_wider(names_from = endemicity, values_from = n, values_fill = 0L) %>%
    mutate(seasonality_signal = factor(seasonality_signal,
      levels = c("weak", "moderate", "strong"))) %>%
    arrange(seasonality_signal) %>%
    knitr::kable()
}
```

```{r}
#| label: fig-strata-scatter
#| fig-cap: "Each country by burden (mean monthly cases, log scale) and seasonal concentration, coloured by provisional endemicity. Dashed lines mark the endemicity case threshold and the seasonality-signal tertile cuts."
#| fig-height: 5

if (!is.null(summary_country)) {
  cuts <- quantile(summary_country$seasonal_concentration, c(1/3, 2/3), na.rm = TRUE)
  ggplot(summary_country,
         aes(mean_monthly_cases, seasonal_concentration, colour = endemicity)) +
    geom_hline(yintercept = cuts, linetype = "dashed", colour = "grey60") +
    geom_vline(xintercept = endemic_min_mean_cases, linetype = "dashed", colour = "grey60") +
    geom_point(size = 2, alpha = 0.8) +
    scale_x_log10(labels = scales::label_comma()) +
    scale_colour_manual(values = c(endemic = "#C44E52", emerging = "#4C72B0")) +
    labs(x = "Mean monthly cases (log scale)", y = "Seasonal concentration (HHI)",
         colour = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-region-endemicity
#| fig-cap: "Countries per region, split by provisional endemicity."
#| fig-height: 3.6

if (!is.null(summary_country)) {
  summary_country %>%
    count(region, endemicity) %>%
    ggplot(aes(n, reorder(region, n, sum), fill = endemicity)) +
    geom_col() +
    scale_fill_manual(values = c(endemic = "#C44E52", emerging = "#4C72B0")) +
    labs(x = "Countries", y = NULL, fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: tbl-region-summary
#| tbl-cap: "Region summary."

if (!is.null(summary_region)) {
  summary_region %>%
    transmute(
      region, n_countries,
      endemic = n_endemic, emerging = n_emerging,
      seasons_median, total_cases = scales::label_comma()(total_cases)
    ) %>%
    knitr::kable()
}
```

::: {.callout-note title="Notes / observations"}
_Add: does the strata split look epidemiologically sensible? Any region with too
few countries in a stratum to score separately? Decision on tertile vs cluster
approach._
:::

---

# Backtest window coverage {#sec-coverage}

`03_check_window_coverage.R` is the pre-Stage-1 gate: for each monthly rolling
origin and each candidate training window (fixed 5 / 7 / 10 yr, expanding), how
many countries have at least `r min_train_months` valued months in the window
**and** at least one target month with truth. No models are fitted.

```{r}
#| label: load-coverage

coverage_country <- read_fc("cv_splits", "rolling_origin_country.csv")

win_lvls <- c("fixed_5y", "fixed_7y", "fixed_10y", "expanding")
if (!is.null(coverage_grid)) {
  coverage_grid <- coverage_grid %>%
    mutate(
      origin_date = as.Date(origin_date),
      window_type = factor(window_type, levels = win_lvls)
    )
}
```

```{r}
#| label: fig-coverage-lines
#| fig-cap: "Countries eligible for a fit at each rolling origin, by training window. Shorter windows lose the countries with gappy recent history; all windows fall away at the final origins as the countries with no 2026 data drop out."
#| fig-height: 3.8

if (is.null(coverage_grid)) {
  pending("Scripts/forecasting/03_check_window_coverage.R")
} else {
  ggplot(coverage_grid, aes(origin_date, n_countries_eligible, colour = window_type)) +
    geom_line(linewidth = 0.7) +
    geom_hline(yintercept = unique(coverage_grid$n_countries_total),
               linetype = "dotted", colour = "grey50") +
    scale_colour_brewer(palette = "Set2") +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    labs(x = NULL, y = "Countries eligible", colour = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: tbl-coverage-summary
#| tbl-cap: "Eligible countries per window, and the first origin reaching 60."

if (!is.null(coverage_grid)) {
  coverage_grid %>%
    arrange(window_type, origin_date) %>%
    group_by(window_type) %>%
    summarise(
      first_origin  = first(n_countries_eligible),
      median_origin = round(median(n_countries_eligible)),
      last_origin   = last(n_countries_eligible),
      reaches_60    = {
        hit <- origin_date[n_countries_eligible >= 60]
        if (length(hit) == 0) "never" else format(min(hit))
      },
      .groups = "drop"
    ) %>%
    knitr::kable()
}
```

```{r}
#| label: coverage-never-eligible
#| output: asis

if (!is.null(coverage_country)) {
  cc <- coverage_country %>%
    mutate(window_type = factor(window_type, levels = win_lvls))
  all_iso <- sort(unique(cc$iso3))

  cat("**Countries never eligible at any origin** (short or gap-dominated records):\n\n")
  for (w in win_lvls) {
    ever <- cc %>% filter(window_type == w, eligible) %>% distinct(iso3) %>% pull(iso3)
    never <- setdiff(all_iso, ever)
    cat(sprintf("- `%s` (%d): %s\n", w, length(never),
                if (length(never)) paste0("`", paste(sort(never), collapse = "`, `"), "`") else "_none_"))
  }
}
```

::: {.callout-tip title="Window viability, by region"}

Median number of origins (of `r if (!is.null(coverage_grid)) length(unique(coverage_grid$origin_date)) else 121`)
at which the typical country in a region is eligible, per window. `121` = every
origin.

```{r}
#| label: tbl-window-viability-region

if (is.null(coverage_country)) {
  pending("Scripts/forecasting/03_check_window_coverage.R")
} else {
  guidance <- tibble::tribble(
    ~region,                              ~guidance,
    "North & Central America",            "any window; modelling choice only",
    "South America",                      "any window (only CHL needs expanding)",
    "South Asia",                         "any window",
    "East & Southeast Asia",              "any window; recent drop = no-2026-data countries, not window length",
    "Pacific Islands",                    "7 yr minimum; 10 yr / expanding covers all 15",
    "Caribbean",                          "7 yr minimum; 5 yr drops ~8 islands at recent origins",
    "Sub-Saharan Africa",                 "expanding only; short windows unusable, several still fall to Baseline 2",
    "Europe, Middle East & North Africa", "no standalone stratum under any window; pool to global"
  )

  n_ctry <- coverage_country %>% distinct(region, iso3) %>% count(region, name = "n")

  coverage_country %>%
    mutate(window_type = factor(window_type, levels = win_lvls)) %>%
    group_by(region, window_type, iso3) %>%
    summarise(n_elig = sum(eligible), .groups = "drop") %>%
    group_by(region, window_type) %>%
    summarise(median_origins = round(median(n_elig)), .groups = "drop") %>%
    pivot_wider(names_from = window_type, values_from = median_origins) %>%
    left_join(n_ctry, by = "region") %>%
    left_join(guidance, by = "region") %>%
    mutate(region = factor(region, levels = region_levels)) %>%
    arrange(region) %>%
    transmute(
      region = paste0(as.character(region), " (", n, ")"),
      `5 yr` = fixed_5y, `7 yr` = fixed_7y, `10 yr` = fixed_10y, expanding,
      guidance
    ) %>%
    knitr::kable(align = c("l", "r", "r", "r", "r", "l"))
}
```

**Default: expanding** — the only window viable everywhere it physically can be
(94 / 96 countries eligible at some origin). **`fixed_10y`** carried as the
trimming comparison for the data-rich regions (≈ expanding for coverage, diverges
only where >10 yr of history exists). **`fixed_7y`** as a Caribbean / Pacific
fallback. **`fixed_5y`** diagnostic only — it cannot be a global default.

:::

::: {.callout-note title="Notes / observations"}
_Add: final window decision, how the chronically-intermittent countries (SSA,
Caribbean tail) are handled, whether fixed-5yr is worth the compute as a
diagnostic._
:::

---

# Stage 0 — model fit diagnostics {#sec-stage0}

`run_stage0_fit.R` fits each model once on the full panel and runs its diagnostic
checklist (convergence, dispersion, residual structure). Diagnostic only — no
skill claim.

```{r}
#| label: tbl-stage0
#| tbl-cap: "Stage 0 fit diagnostics."

if (is.null(stage0_diag)) {
  pending("Scripts/forecasting/run_stage0_fit.R")
} else {
  knitr::kable(stage0_diag)
}
```

::: {.callout-note title="Notes / observations"}
_Add: did every candidate pass the sanity check? Any coefficient signs or
dispersion values worth flagging before Stage 1._
:::

---

# Stage 1 — hindcast skill {#sec-stage1}

Rolling-origin backtest across the four window lengths and the scenario strata.
Primary metrics: WIS (log scale) and peak-timing difference. Secondary: MAE,
RMSE, CRPS, PIT / coverage.

```{r}
#| label: sec-stage1-body

pending("the Stage 1 backtest runner (not built yet)")
```

Planned outputs to land here: relative-skill leaderboard by lead time, score
table (model × lead × stratum, with `n_seasons`), score heatmap, per-country fan
charts, PIT / coverage calibration.

::: {.callout-note title="Notes / observations"}
_Add as results arrive._
:::

---

# Conclusions & manuscript notes {#sec-conclusions}

Running synthesis for the forecasting paper (target: PLOS Comp Biol / PLOS NTD).

- **Data.** _..._
- **Strata.** _..._
- **Baselines.** _..._
- **Best model(s) so far.** _..._
- **Where the forecast is trustworthy / where it is not.** _..._

---

# Open questions & directions {#sec-directions}

- Minimum complete-seasons floor below which a country defaults to the seasonal
  baseline (`[Q2.4]`).
- Spatial (region-grouped) blocking alongside the temporal windows (`§6`).
- Promotion rule: beat both baselines at *majority* vs *all* lead times (`§6`).
- Refit cadence for the hierarchical model — monthly vs quarterly (`[Q2.6]`).
- Operational-utility scoring design at ~83-country scale (`[Q5.2]`).
- Climate covariate sources and the GAM / hierarchical model structure.

---

# Provenance {.unnumbered}

```{r}
#| label: provenance
#| echo: false

tibble::tibble(
  item = c("Data snapshot", "Forecast horizons", "Rolling windows (months)",
           "Min training months", "Origin start", "Rendered"),
  value = c(
    snapshot_date,
    paste(range(forecast_horizons), collapse = "–"),
    paste(names(roll_windows), unname(roll_windows), sep = "=", collapse = ", "),
    as.character(min_train_months),
    format(origin_start),
    format(Sys.time(), "%Y-%m-%d %H:%M")
  )
) %>% knitr::kable()
```

```{r}
#| label: sessioninfo
#| code-fold: true
#| code-summary: "R session"

sessionInfo()
```
