---
title: "GDO Forecasting — Evaluation Notebook"
subtitle: "Results, figures and working notes for the forecasting objective"
author: "K M Susong"
date: today
format:
  html:
    grid:
      body-width: 1000px
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(lubridate)
library(ggplot2)
library(knitr)

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 4.5, dpi = 150
)

theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold")))

# --- Locate the repo root, whatever directory this renders from ---
find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/")
  repeat {
    if (file.exists(file.path(d, "Scripts", "V1_Pipeline.R"))) return(d)
    parent <- dirname(d)
    if (identical(parent, d)) {
      stop("DENV_global_observatory repo root not found from ", start)
    }
    d <- parent
  }
}
repo_root <- find_repo_root()

# Pinned parameters and provisional thresholds come straight from the pipeline
# config, so this notebook always reports what the scripts actually used.
source(file.path(repo_root, "Scripts", "forecasting", "00_config.R"), local = TRUE)

fc_out <- file.path(repo_root, forecast_out)

# Load a forecasting output if it exists yet, else NULL.
read_fc <- function(...) {
  p <- file.path(fc_out, ...)
  if (!file.exists(p)) return(NULL)
  readr::read_csv(p, show_col_types = FALSE)
}

# Emit a "not generated yet" callout for sections whose script has not run.
# asis_output renders correctly whether or not the calling chunk is `asis`.
pending <- function(script) {
  knitr::asis_output(sprintf(
    "::: {.callout-note appearance=\"simple\"}\nNot generated yet -- run `%s` and re-render.\n:::\n",
    script
  ))
}
```

```{r}
#| label: load-outputs

panel          <- read_fc("training_data", paste0("training_panel_", snapshot_date, ".csv"))
region_lookup  <- read_fc("training_data", "country_region_lookup.csv")
summary_country <- read_fc("data_summary", "summary_country.csv")
summary_region  <- read_fc("data_summary", "summary_region.csv")
strata_country  <- read_fc("data_summary", "strata_country.csv")
coverage_grid   <- read_fc("cv_splits", "rolling_origins.csv")
stage0_diag     <- read_fc("stage0_fit", "diagnostics.csv")

if (!is.null(panel)) {
  panel <- panel %>%
    mutate(
      date = as.Date(date),
      data_status = factor(data_status,
        levels = c("observed", "corrected", "estimated", "gap"))
    )
}
```

## About this notebook {.unnumbered}

This is the **results record** for the GDO forecasting objective — the place to
look at what each pipeline step produced, keep the figures and tables that will
go into the manuscript, and write down what they mean and where the work goes
next.

It is deliberately **not**:

- the running task log — that lives in Logseq;
- a home for pipeline code — the training-data build, data summary, model
  definitions, runners and scoring pipeline are all under
  `Scripts/forecasting/`. This notebook only *reads their outputs*.

::: {.callout-note appearance="simple"}
**Pinned data snapshot:** `Output/`r snapshot_date`/` — merged historical record
plus current-period nowcast output. Every result below is against this frozen
snapshot; re-pin in `Scripts/forecasting/00_config.R` and re-run the pipeline to
refresh.
:::

Plan of record: `planning/2026_forecast_evaluation/03_Evaluation_Plan_DRAFT`.
Stages: **0** in-sample fit diagnostics · **1** rolling-origin hindcast
(5 / 7 / 10 yr + expanding) · **2** prospective shadow deployment.

---

# Training data {#sec-training-data}

`01_prepare_training_data.R` merges the deep historical record
(`full_data_season_monthly_proportions.csv`, complete seasons) with the
current-period nowcast output (`DENV_cases_nowcast_output.csv`). The nowcast view
wins on any overlapping country-month, so the panel runs as far forward as data
exists and carries the post-nowcast "ground truth".

`data_status` is preserved throughout: **observed** (raw report), **corrected**
(reporting-delay factor applied), **estimated** (seasonal-proportion nowcast),
**gap** (internal month with no value — the historical file drops all-zero
years).

```{r}
#| label: tbl-panel-overview
#| tbl-cap: "Training panel — coverage and composition."

if (is.null(panel)) {
  pending("Scripts/forecasting/01_prepare_training_data.R")
} else {
  tibble::tibble(
    Metric = c("Countries", "Country-months", "Date range",
               "From nowcast output", "From history"),
    Value = c(
      as.character(n_distinct(panel$iso3)),
      as.character(nrow(panel)),
      paste(format(min(panel$date)), "to", format(max(panel$date))),
      "", ""
    )
  ) %>% knitr::kable()
}
```

```{r}
#| label: tbl-data-status
#| tbl-cap: "Country-months by data status."

if (!is.null(panel)) {
  panel %>%
    count(data_status, name = "country_months") %>%
    mutate(share = scales::percent(country_months / sum(country_months), 0.1)) %>%
    knitr::kable()
}
```

```{r}
#| label: fig-data-status-year
#| fig-cap: "Data-status composition by calendar year. Recent years carry the reporting-delay corrections and seasonal-proportion estimates the forecast layer treats as ground truth."
#| fig-height: 4

if (!is.null(panel)) {
  panel %>%
    filter(Year >= 2018) %>%
    count(Year, data_status) %>%
    ggplot(aes(Year, n, fill = data_status)) +
    geom_col() +
    scale_fill_manual(values = c(
      observed = "#4C72B0", corrected = "#DD8452",
      estimated = "#C44E52", gap = "grey80"
    )) +
    labs(x = NULL, y = "Country-months", fill = "Status") +
    theme(legend.position = "top")
}
```

```{r}
#| label: tbl-country-coverage
#| tbl-cap: "Per-country coverage (complete seasons, span, and share of months that are model-estimated). Full detail in summary_country.csv."

if (is.null(summary_country)) {
  pending("Scripts/forecasting/02_data_summary.R")
} else {
  summary_country %>%
    transmute(
      iso3, country, region,
      complete_seasons = n_complete_seasons,
      months = n_months,
      span_yrs = round(span_years, 1),
      gap_months = n_gap_months,
      pct_estimated = scales::percent(pct_estimated, 0.1)
    ) %>%
    arrange(region, desc(complete_seasons))
}
```

::: {.callout-note title="Notes / observations"}
_Add: which countries are thin on history, how much of the recent record is
corrected vs estimated, anything that will constrain the backtest windows._
:::

---

# Data characterisation & scenario strata {#sec-strata}

`02_data_summary.R` characterises each country's burden and seasonal shape, and
assigns the two **provisional** scenario strata the Stage 1 backtest will be
sliced on (the third axis, lead time, is intrinsic to the forecast).

Seasonal shape is read from the mean seasonal profile (`DENV_average_season.csv`):
`seasonal_concentration` is the Herfindahl index of the mean monthly proportions
— `r round(1/12, 3)` for a flat year, rising toward 1 as a single month
dominates.

::: {.callout-important appearance="simple"}
**Provisional cut-points — to confirm with the team.**

- **seasonality_signal**: tertiles of `seasonal_concentration` across the
  `r if (!is.null(summary_country)) nrow(summary_country) else 96` countries →
  weak / moderate / strong.
- **endemicity**: *endemic* only if mean monthly cases ≥
  `r endemic_min_mean_cases`, complete seasons ≥ `r endemic_min_seasons`, and
  zero-month fraction < `r endemic_max_zero_fraction`; otherwise *emerging*.

Alternative under consideration: adopt the seasonality clusters from Joshi et al.
in place of the tertile split.
:::

```{r}
#| label: tbl-strata-crosstab
#| tbl-cap: "Countries by provisional stratum."

if (is.null(summary_country)) {
  pending("Scripts/forecasting/02_data_summary.R")
} else {
  summary_country %>%
    count(seasonality_signal, endemicity) %>%
    pivot_wider(names_from = endemicity, values_from = n, values_fill = 0L) %>%
    mutate(seasonality_signal = factor(seasonality_signal,
      levels = c("weak", "moderate", "strong"))) %>%
    arrange(seasonality_signal) %>%
    knitr::kable()
}
```

```{r}
#| label: fig-strata-scatter
#| fig-cap: "Each country by burden (mean monthly cases, log scale) and seasonal concentration, coloured by provisional endemicity. Dashed lines mark the endemicity case threshold and the seasonality-signal tertile cuts."
#| fig-height: 5

if (!is.null(summary_country)) {
  cuts <- quantile(summary_country$seasonal_concentration, c(1/3, 2/3), na.rm = TRUE)
  ggplot(summary_country,
         aes(mean_monthly_cases, seasonal_concentration, colour = endemicity)) +
    geom_hline(yintercept = cuts, linetype = "dashed", colour = "grey60") +
    geom_vline(xintercept = endemic_min_mean_cases, linetype = "dashed", colour = "grey60") +
    geom_point(size = 2, alpha = 0.8) +
    scale_x_log10(labels = scales::label_comma()) +
    scale_colour_manual(values = c(endemic = "#C44E52", emerging = "#4C72B0")) +
    labs(x = "Mean monthly cases (log scale)", y = "Seasonal concentration (HHI)",
         colour = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-region-endemicity
#| fig-cap: "Countries per region, split by provisional endemicity."
#| fig-height: 3.6

if (!is.null(summary_country)) {
  summary_country %>%
    count(region, endemicity) %>%
    ggplot(aes(n, reorder(region, n, sum), fill = endemicity)) +
    geom_col() +
    scale_fill_manual(values = c(endemic = "#C44E52", emerging = "#4C72B0")) +
    labs(x = "Countries", y = NULL, fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: tbl-region-summary
#| tbl-cap: "Region summary."

if (!is.null(summary_region)) {
  summary_region %>%
    transmute(
      region, n_countries,
      endemic = n_endemic, emerging = n_emerging,
      seasons_median, total_cases = scales::label_comma()(total_cases)
    ) %>%
    knitr::kable()
}
```

::: {.callout-note title="Notes / observations"}
_Add: does the strata split look epidemiologically sensible? Any region with too
few countries in a stratum to score separately? Decision on tertile vs cluster
approach._
:::

---

# Backtest window coverage {#sec-coverage}

`03_check_window_coverage.R` — the pre-Stage-1 gate: for each rolling origin and
each window length (5 / 7 / 10 yr / expanding), how many countries have enough
history to enter a fit.

```{r}
#| label: sec-coverage-body

if (is.null(coverage_grid)) {
  pending("Scripts/forecasting/03_check_window_coverage.R")
} else {
  knitr::asis_output("_(coverage tables and figures to be added once the script lands)_\n")
}
```

::: {.callout-note title="Notes / observations"}
_Add: earliest usable origin per window, which countries never clear the 10-year
window, whether any window is too thin to bother testing._
:::

---

# Stage 0 — model fit diagnostics {#sec-stage0}

`run_stage0_fit.R` fits each model once on the full panel and runs its diagnostic
checklist (convergence, dispersion, residual structure). Diagnostic only — no
skill claim.

```{r}
#| label: tbl-stage0
#| tbl-cap: "Stage 0 fit diagnostics."

if (is.null(stage0_diag)) {
  pending("Scripts/forecasting/run_stage0_fit.R")
} else {
  knitr::kable(stage0_diag)
}
```

::: {.callout-note title="Notes / observations"}
_Add: did every candidate pass the sanity check? Any coefficient signs or
dispersion values worth flagging before Stage 1._
:::

---

# Stage 1 — hindcast skill {#sec-stage1}

Rolling-origin backtest across the four window lengths and the scenario strata.
Primary metrics: WIS (log scale) and peak-timing difference. Secondary: MAE,
RMSE, CRPS, PIT / coverage.

```{r}
#| label: sec-stage1-body

pending("the Stage 1 backtest runner (not built yet)")
```

Planned outputs to land here: relative-skill leaderboard by lead time, score
table (model × lead × stratum, with `n_seasons`), score heatmap, per-country fan
charts, PIT / coverage calibration.

::: {.callout-note title="Notes / observations"}
_Add as results arrive._
:::

---

# Conclusions & manuscript notes {#sec-conclusions}

Running synthesis for the forecasting paper (target: PLOS Comp Biol / PLOS NTD).

- **Data.** _..._
- **Strata.** _..._
- **Baselines.** _..._
- **Best model(s) so far.** _..._
- **Where the forecast is trustworthy / where it is not.** _..._

---

# Open questions & directions {#sec-directions}

- Minimum complete-seasons floor below which a country defaults to the seasonal
  baseline (`[Q2.4]`).
- Spatial (region-grouped) blocking alongside the temporal windows (`§6`).
- Promotion rule: beat both baselines at *majority* vs *all* lead times (`§6`).
- Refit cadence for the hierarchical model — monthly vs quarterly (`[Q2.6]`).
- Operational-utility scoring design at ~83-country scale (`[Q5.2]`).
- Climate covariate sources and the GAM / hierarchical model structure.

---

# Provenance {.unnumbered}

```{r}
#| label: provenance
#| echo: false

tibble::tibble(
  item = c("Data snapshot", "Forecast horizons", "Rolling windows (months)",
           "Min training months", "Origin start", "Rendered"),
  value = c(
    snapshot_date,
    paste(range(forecast_horizons), collapse = "–"),
    paste(names(roll_windows), unname(roll_windows), sep = "=", collapse = ", "),
    as.character(min_train_months),
    format(origin_start),
    format(Sys.time(), "%Y-%m-%d %H:%M")
  )
) %>% knitr::kable()
```

```{r}
#| label: sessioninfo
#| code-fold: true
#| code-summary: "R session"

sessionInfo()
```
