---
title: "GDO Forecasting — Evaluation Notebook"
subtitle: "Results, figures and working notes for the forecasting objective"
author: "K M Susong"
date: today
format:
  html:
    grid:
      body-width: 1000px
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(lubridate)
library(ggplot2)
library(knitr)

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 4.5, dpi = 150
)

theme_set(theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold")))

# --- Locate the repo root, whatever directory this renders from ---
find_repo_root <- function(start = getwd()) {
  d <- normalizePath(start, winslash = "/")
  repeat {
    if (file.exists(file.path(d, "Scripts", "V1_Pipeline.R"))) return(d)
    parent <- dirname(d)
    if (identical(parent, d)) {
      stop("DENV_global_observatory repo root not found from ", start)
    }
    d <- parent
  }
}
repo_root <- find_repo_root()

# Pinned parameters and provisional thresholds come straight from the pipeline
# config, so this notebook always reports what the scripts actually used.
source(file.path(repo_root, "Scripts", "forecasting", "00_config.R"), local = TRUE)

fc_out <- file.path(repo_root, forecast_out)

# Load a forecasting output if it exists yet, else NULL.
read_fc <- function(...) {
  p <- file.path(fc_out, ...)
  if (!file.exists(p)) return(NULL)
  readr::read_csv(p, show_col_types = FALSE)
}

# Emit a "not generated yet" callout for sections whose script has not run.
# asis_output renders correctly whether or not the calling chunk is `asis`.
pending <- function(script) {
  knitr::asis_output(sprintf(
    "::: {.callout-note appearance=\"simple\"}\nNot generated yet -- run `%s` and re-render.\n:::\n",
    script
  ))
}
```

```{r}
#| label: load-outputs

panel          <- read_fc("training_data", paste0("training_panel_", snapshot_date, ".csv"))
region_lookup  <- read_fc("training_data", "country_region_lookup.csv")
summary_country <- read_fc("data_summary", "summary_country.csv")
summary_region  <- read_fc("data_summary", "summary_region.csv")
strata_country  <- read_fc("data_summary", "strata_country.csv")
coverage_grid   <- read_fc("cv_splits", "rolling_origins.csv")
stage0_diag     <- read_fc("stage0_fit", "diagnostics.csv")

if (!is.null(panel)) {
  panel <- panel %>%
    mutate(
      date = as.Date(date),
      data_status = factor(data_status,
        levels = c("observed", "corrected", "estimated", "gap"))
    )
}
```

## About this notebook {.unnumbered}

This is the **results record** for the GDO forecasting objective — the place to look at what each pipeline step produced, keep the figures and tables that will go into the manuscript, and write down what they mean and where the work goes next.

It is deliberately **not**:

- the running task log — that lives in Logseq;
- a home for pipeline code — the training-data build, data summary, model definitions, runners and scoring pipeline are all under `Scripts/forecasting/`. This notebook only *reads their outputs*.

::: {.callout-note appearance="simple"}
**Pinned data snapshot:** `Output/`r snapshot_date`/` — merged historical record plus current-period nowcast output. Every result below is against this frozen snapshot; re-pin in `Scripts/forecasting/00_config.R` and re-run the pipeline to refresh.
:::

Plan of record: `planning/2026_forecast_evaluation/03_Evaluation_Plan_DRAFT`. Stages: **0** in-sample fit diagnostics · **1** rolling-origin hindcast (5 / 7 / 10 yr + expanding) · **2** prospective shadow deployment.

------------------------------------------------------------------------

# Training data {#sec-training-data}

`01_prepare_training_data.R` merges the deep historical record (`full_data_season_monthly_proportions.csv`, complete seasons) with the current-period nowcast output (`DENV_cases_nowcast_output.csv`). The nowcast view wins on any overlapping country-month, so the panel runs as far forward as data exists and carries the post-nowcast "ground truth".

`data_status` is preserved throughout: **observed** (raw report), **corrected** (reporting-delay factor applied), **estimated** (seasonal-proportion nowcast), **gap** (internal month with no value — the historical file drops all-zero years).

```{r}
#| label: status-palette

status_pal <- c(
  observed  = "#4C72B0",
  corrected = "#DD8452",
  estimated = "#C44E52",
  gap       = "grey85"
)

if (!is.null(panel)) {
  panel <- panel %>%
    mutate(region = factor(region, levels = region_levels))
}
```

```{r}
#| label: tbl-panel-overview
#| tbl-cap: "Training panel: coverage and composition."

if (is.null(panel)) {
  pending("Scripts/forecasting/01_prepare_training_data.R")
} else {
  fmt <- function(x) formatC(x, big.mark = ",", format = "d")
  tibble::tibble(
    Metric = c("Countries", "Country-months", "Date range",
               "Observed", "Corrected (reporting-delay factor)",
               "Estimated (seasonal-proportion nowcast)",
               "Gap (internal month, no value)"),
    Value = c(
      fmt(n_distinct(panel$iso3)),
      fmt(nrow(panel)),
      paste(format(min(panel$date)), "to", format(max(panel$date))),
      fmt(sum(panel$data_status == "observed")),
      fmt(sum(panel$data_status == "corrected")),
      fmt(sum(panel$data_status == "estimated")),
      fmt(sum(panel$data_status == "gap"))
    )
  ) %>% knitr::kable(align = c("l", "r"))
}
```

```{r}
#| label: fig-data-status-year
#| fig-cap: "Data-status composition by calendar year. Recent years carry the reporting-delay corrections and seasonal-proportion estimates the forecast layer treats as ground truth."
#| fig-height: 3.6

if (!is.null(panel)) {
  panel %>%
    filter(Year >= 2018) %>%
    count(Year, data_status) %>%
    ggplot(aes(Year, n, fill = data_status)) +
    geom_col() +
    scale_fill_manual(values = status_pal) +
    labs(x = NULL, y = "Country-months", fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-gap-tiles
#| fig-cap: "Gap months per country-year (months the historical record drops as all-zero, shown as missing for the forecast layer). Blank cells are outside a country's covered span. Faceted by region."
#| fig-width: 8.5
#| fig-height: 12
#| column: body-outset

if (!is.null(panel)) {
  panel %>%
    group_by(region, iso3, Year) %>%
    summarise(gap_months = sum(data_status == "gap"), .groups = "drop") %>%
    ggplot(aes(Year, iso3, fill = gap_months)) +
    geom_tile(colour = "grey40", linewidth = 0.15) +
    facet_grid(region ~ ., scales = "free_y", space = "free_y") +
    scale_fill_gradient(low = "white", high = "#C44E52",
                        breaks = c(0, 3, 6, 9, 12), limits = c(0, 12)) +
    scale_x_continuous(breaks = seq(2010, 2026, 2), expand = c(0, 0)) +
    labs(x = NULL, y = NULL, fill = "Gap months",
         caption = "Grey background = year outside the country's record.") +
    theme(
      axis.text.y = element_text(size = 6),
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "grey82", colour = NA),
      strip.text.y = element_text(angle = 0, hjust = 0, size = 8),
      legend.position = "top"
    )
}
```

```{r}
#| label: fig-status-tiles
#| fig-cap: "Monthly data status for every country, 2010 to present. The forecast layer trains on observed + corrected + estimated months; gaps are internal missing months. Faceted by region."
#| fig-width: 9
#| fig-height: 13
#| column: body-outset

if (!is.null(panel)) {
  panel %>%
    ggplot(aes(date, iso3, fill = data_status)) +
    geom_tile() +
    facet_grid(region ~ ., scales = "free_y", space = "free_y") +
    scale_fill_manual(values = status_pal, drop = FALSE) +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y", expand = c(0, 0)) +
    labs(x = NULL, y = NULL, fill = NULL) +
    theme(
      axis.text.y = element_text(size = 6),
      panel.grid = element_blank(),
      strip.text.y = element_text(angle = 0, hjust = 0, size = 8),
      legend.position = "top"
    )
}
```

```{r}
#| label: tbl-country-coverage
#| tbl-cap: "Per-country coverage, searchable reference (full detail in summary_country.csv)."

if (is.null(summary_country)) {
  pending("Scripts/forecasting/02_data_summary.R")
} else {
  cov_tbl <- summary_country %>%
    transmute(
      iso3, country, region,
      complete_seasons = n_complete_seasons,
      months = n_months,
      span_yrs = round(span_years, 1),
      gap_months = n_gap_months,
      pct_estimated = round(100 * pct_estimated, 1)
    ) %>%
    arrange(region, desc(complete_seasons))

  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(
      cov_tbl,
      rownames = FALSE,
      filter = "top",
      options = list(pageLength = 15, order = list(list(2, "asc")))
    )
  } else {
    knitr::kable(cov_tbl)
  }
}
```

::: {.callout-warning title="Coverage flags — countries that will constrain the backtest"}
```{r}
#| label: coverage-flags
#| output: asis

if (is.null(panel) || is.null(summary_country)) {
  pending("Scripts/forecasting/01_prepare_training_data.R and 02_data_summary.R")
} else {
  fmt_iso <- function(x) {
    if (length(x) == 0) return("_none_")
    paste0("`", paste(sort(x), collapse = "`, `"), "`")
  }

  # 1. No observed / corrected / estimated month in 2026.
  in_2026 <- panel %>%
    filter(Year == 2026, data_status != "gap") %>%
    distinct(iso3) %>%
    pull(iso3)
  no_2026 <- setdiff(unique(panel$iso3), in_2026)

  # 2. Fewer than 60 months (5 years) of actual data, whatever the span.
  thin <- summary_country %>%
    filter(n_months < 60) %>%
    arrange(n_months)

  # 3. Gap months inside the last 5 years.
  cutoff <- max(panel$date) %m-% years(5)
  gaps_5y <- panel %>%
    filter(date >= cutoff, data_status == "gap") %>%
    count(iso3, name = "gap_months") %>%
    arrange(desc(gap_months))
  heavy <- gaps_5y %>% filter(gap_months >= 12)
  light <- gaps_5y %>% filter(gap_months < 12)

  cat(sprintf(
    "**No data in 2026** (%d) -- seasonal profile only, no current-season observation; carried for history but not available for prospective scoring: %s\n\n",
    length(no_2026), fmt_iso(no_2026)
  ))

  cat(sprintf(
    "**Under 5 years of actual data** (%d) -- too thin for a fixed 5-year window; candidates to default to the seasonal baseline:\n\n",
    nrow(thin)
  ))
  cat("| iso3 | months | complete seasons | span (yr) |\n")
  cat("|------|-------:|-----------------:|----------:|\n")
  cat(paste(sprintf(
    "| `%s` | %d | %d | %.1f |",
    thin$iso3, thin$n_months, thin$n_complete_seasons, thin$span_years
  ), collapse = "\n"))
  cat("\n\n")

  cat(sprintf(
    "**Gaps in the last 5 years** (%d countries, since %s) -- a fixed-window backtest trains on discontinuous history here:\n\n",
    nrow(gaps_5y), format(cutoff)
  ))
  cat(sprintf(
    "- 12+ gap months (%d): %s\n\n",
    nrow(heavy),
    paste(sprintf("`%s` (%d)", heavy$iso3, heavy$gap_months), collapse = ", ")
  ))
  cat(sprintf("- fewer than 12 (%d): %s\n", nrow(light), fmt_iso(light$iso3)))
}
```
:::

::: {.callout-note title="Notes / observations"}
*Add: reading of the flags above — which of these default to the seasonal baseline vs still get a bespoke model, whether the expanding window rescues the gappy countries, how much of the recent record is corrected vs estimated.*
:::

------------------------------------------------------------------------

# Covariate & climate data {#sec-covariates}

Placeholder for climate and other covariate data once sourced (plan §1, item 4 — GLM+climate / GAM+climate). Candidate variables, sources, coverage by country/region, and any exploratory relationship with case counts will go here once that sourcing work starts.

::: {.callout-note appearance="simple"}
Not sourced yet — no covariate data exists in the repo at this point.
:::

::: {.callout-note title="Notes / observations"}
*Add: candidate covariates under consideration, sourcing plan, coverage.*
:::

------------------------------------------------------------------------

# Data characterisation & scenario strata {#sec-strata}

`02_data_summary.R` characterises each country's burden, seasonal shape and land area, and assigns the **provisional** scenario strata the Stage 1 hindcast will be sliced on (the remaining axis, lead time, is intrinsic to the forecast).

Seasonal shape is read from the mean seasonal profile (`DENV_average_season.csv`): `seasonal_concentration` is the Herfindahl index of the mean monthly proportions — `r round(1/12, 3)` for a flat year, rising toward 1 as a single month dominates. Land area is computed from the country outline geometry (`base_map.gpkg`), equal-area projected.

::: {.callout-important appearance="simple"}
**Provisional cut-points — to confirm with the team.**

- **seasonality_signal**: tertiles of `seasonal_concentration` across the `r if (!is.null(summary_country)) nrow(summary_country) else 96` countries → weak / moderate / strong.
- **endemicity**: *endemic* only if mean monthly cases ≥ `r endemic_min_mean_cases`, complete seasons ≥ `r endemic_min_seasons`, and zero-month fraction \< `r endemic_max_zero_fraction`; otherwise *emerging*.
- **size_class**: land area `cut()` on fixed km² breaks — `r if (exists("size_labels")) paste(size_labels, collapse = " / ") else "very_small / small / medium / large / continental"`. Change the breaks or the class count in `00_config.R`.

Alternative under consideration for seasonality: adopt Kishen Joshi's seasonality clusters in place of the tertile split (ask him for the file).
:::

```{r}
#| label: strata-crosstab-helper

# A 2-way count with a Total column and a Total row, for the small strata
# crosstabs below. `row_var` stays a factor so the row order is kept.
strata_crosstab <- function(df, row_var, col_var) {
  wide <- df %>%
    count({{ row_var }}, {{ col_var }}, .drop = FALSE) %>%
    pivot_wider(names_from = {{ col_var }}, values_from = n, values_fill = 0L) %>%
    arrange({{ row_var }}) %>%
    mutate(Total = rowSums(across(where(is.numeric))))
  total_row <- wide %>%
    summarise(across(where(is.numeric), sum)) %>%
    mutate("{{ row_var }}" := "Total", .before = 1)
  bind_rows(mutate(wide, "{{ row_var }}" := as.character({{ row_var }})), total_row)
}
```

```{r}
#| label: fig-region-seasonality
#| fig-cap: "Countries per region, split by seasonality signal (weak -> strong)."
#| fig-height: 3.6

if (!is.null(summary_country)) {
  summary_country %>%
    mutate(seasonality_signal = factor(seasonality_signal,
      levels = c("weak", "moderate", "strong"))) %>%
    count(region, seasonality_signal, .drop = FALSE) %>%
    ggplot(aes(n, reorder(region, n, sum), fill = seasonality_signal)) +
    geom_col() +
    scale_fill_brewer(palette = "YlOrRd", direction = 1) +
    labs(x = "Countries", y = NULL, fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-region-endemicity
#| fig-cap: "Countries per region, split by provisional endemicity."
#| fig-height: 3.6

if (!is.null(summary_country)) {
  summary_country %>%
    count(region, endemicity) %>%
    ggplot(aes(n, reorder(region, n, sum), fill = endemicity)) +
    geom_col() +
    scale_fill_manual(values = c(endemic = "#C44E52", emerging = "#4C72B0")) +
    labs(x = "Countries", y = NULL, fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: fig-region-size
#| fig-cap: "Countries per region, split by land-area class (very_small -> continental)."
#| fig-height: 3.6

if (!is.null(summary_country)) {
  summary_country %>%
    mutate(size_class = factor(size_class, levels = size_labels)) %>%
    count(region, size_class, .drop = FALSE) %>%
    ggplot(aes(n, reorder(region, n, sum), fill = size_class)) +
    geom_col() +
    scale_fill_viridis_d(option = "mako", direction = -1, end = 0.92) +
    labs(x = "Countries", y = NULL, fill = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: tbl-strata-crosstab
#| tbl-cap: "Countries by provisional stratum: seasonality signal x endemicity."

if (is.null(summary_country)) {
  pending("Scripts/forecasting/02_data_summary.R")
} else {
  summary_country %>%
    mutate(seasonality_signal = factor(seasonality_signal,
      levels = c("weak", "moderate", "strong"))) %>%
    strata_crosstab(seasonality_signal, endemicity) %>%
    knitr::kable()
}
```

```{r}
#| label: tbl-strata-size
#| tbl-cap: "Countries by land-area class (x endemicity). Classes and km2 breaks are set in 00_config.R."

if (!is.null(summary_country)) {
  summary_country %>%
    mutate(size_class = factor(size_class, levels = size_labels)) %>%
    strata_crosstab(size_class, endemicity) %>%
    knitr::kable()
}
```

::: {.callout-note title="Notes / observations"}
**Endemicity**

- *I am surprised that the USA seems to be classified as endemic. The monthly case count min does not account for population, that maybe a problem, should it be a monthly average incidence to account for population size?*

**Seasonality**

- *seasonality tertile will be replaced Kishen's clusters as soon as I have the file*

**Size Class**

- *`size_class` is very_small-heavy (island states) — is the current 5-class / km² cut the right granularity, or collapse to small/medium/large or small/large?*

- *There are region patterns to the distribution on nation size. Will the cause confounding effects to consider?*
:::

```{r}
#| label: tbl-strata-lookup
#| tbl-cap: "Every country's assignment on all three provisional strata, with the underlying values. Searchable; full detail in strata_country.csv."

if (!is.null(strata_country)) {
  lk_tbl <- strata_country %>%
    transmute(
      iso3, country, region,
      seasonality_signal, endemicity, size_class,
      seas_conc      = round(seasonal_concentration, 3),
      mean_cases_mo  = round(mean_monthly_cases),
      seasons        = n_complete_seasons,
      area_km2       = round(country_area_km2)
    ) %>%
    arrange(region, country)

  if (requireNamespace("DT", quietly = TRUE)) {
    DT::datatable(
      lk_tbl,
      rownames = FALSE,
      filter = "top",
      options = list(pageLength = 15, order = list(list(2, "asc")))
    )
  } else {
    knitr::kable(lk_tbl)
  }
}
```

------------------------------------------------------------------------

# Backtest window coverage {#sec-coverage}

`03_check_window_coverage.R` is the pre-Stage-1 gate: for each monthly rolling origin and each candidate training window (fixed 5 / 7 / 10 yr, expanding), how many countries have at least `r min_train_months` valued months in the window **and** at least one target month with truth. No models are fitted.

```{r}
#| label: load-coverage

coverage_country <- read_fc("cv_splits", "rolling_origin_country.csv")

win_lvls <- c("fixed_5y", "fixed_7y", "fixed_10y", "expanding")
if (!is.null(coverage_grid)) {
  coverage_grid <- coverage_grid %>%
    mutate(
      origin_date = as.Date(origin_date),
      window_type = factor(window_type, levels = win_lvls)
    )
}
```

```{r}
#| label: fig-coverage-lines
#| fig-cap: "Countries eligible for a fit at each rolling origin, by training window. Shorter windows lose the countries with gappy recent history; all windows fall away at the final origins as the countries with no 2026 data drop out."
#| fig-height: 3.8

if (is.null(coverage_grid)) {
  pending("Scripts/forecasting/03_check_window_coverage.R")
} else {
  ggplot(coverage_grid, aes(origin_date, n_countries_eligible, colour = window_type)) +
    geom_line(linewidth = 0.7) +
    geom_hline(yintercept = unique(coverage_grid$n_countries_total),
               linetype = "dotted", colour = "grey50") +
    scale_colour_brewer(palette = "Set2") +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    labs(x = NULL, y = "Countries eligible", colour = NULL) +
    theme(legend.position = "top")
}
```

```{r}
#| label: tbl-coverage-summary
#| tbl-cap: "Eligible countries per window, and the first origin reaching 60."

if (!is.null(coverage_grid)) {
  coverage_grid %>%
    arrange(window_type, origin_date) %>%
    group_by(window_type) %>%
    summarise(
      first_origin  = first(n_countries_eligible),
      median_origin = round(median(n_countries_eligible)),
      last_origin   = last(n_countries_eligible),
      reaches_60    = {
        hit <- origin_date[n_countries_eligible >= 60]
        if (length(hit) == 0) "never" else format(min(hit))
      },
      .groups = "drop"
    ) %>%
    knitr::kable()
}
```

```{r}
#| label: coverage-never-eligible
#| output: asis

if (!is.null(coverage_country)) {
  cc <- coverage_country %>%
    mutate(window_type = factor(window_type, levels = win_lvls))
  all_iso <- sort(unique(cc$iso3))

  cat("**Countries never eligible at any origin** (short or gap-dominated records):\n\n")
  for (w in win_lvls) {
    ever <- cc %>% filter(window_type == w, eligible) %>% distinct(iso3) %>% pull(iso3)
    never <- setdiff(all_iso, ever)
    cat(sprintf("- `%s` (%d): %s\n", w, length(never),
                if (length(never)) paste0("`", paste(sort(never), collapse = "`, `"), "`") else "_none_"))
  }
}
```

::: {.callout-tip title="Window viability, by region"}
Median number of origins (of `r if (!is.null(coverage_grid)) length(unique(coverage_grid$origin_date)) else 121`) at which the typical country in a region is eligible, per window. `121` = every origin.

```{r}
#| label: tbl-window-viability-region

if (is.null(coverage_country)) {
  pending("Scripts/forecasting/03_check_window_coverage.R")
} else {
  guidance <- tibble::tribble(
    ~region,                              ~guidance,
    "North & Central America",            "any window; modelling choice only",
    "South America",                      "any window (only CHL needs expanding)",
    "South Asia",                         "any window",
    "East & Southeast Asia",              "any window; recent drop = no-2026-data countries, not window length",
    "Pacific Islands",                    "7 yr minimum; 10 yr / expanding covers all 15",
    "Caribbean",                          "7 yr minimum; 5 yr drops ~8 islands at recent origins",
    "Sub-Saharan Africa",                 "expanding only; short windows unusable, several still fall to Baseline 2",
    "Europe, Middle East & North Africa", "no standalone stratum under any window; pool to global"
  )

  n_ctry <- coverage_country %>% distinct(region, iso3) %>% count(region, name = "n")

  coverage_country %>%
    mutate(window_type = factor(window_type, levels = win_lvls)) %>%
    group_by(region, window_type, iso3) %>%
    summarise(n_elig = sum(eligible), .groups = "drop") %>%
    group_by(region, window_type) %>%
    summarise(median_origins = round(median(n_elig)), .groups = "drop") %>%
    pivot_wider(names_from = window_type, values_from = median_origins) %>%
    left_join(n_ctry, by = "region") %>%
    left_join(guidance, by = "region") %>%
    mutate(region = factor(region, levels = region_levels)) %>%
    arrange(region) %>%
    transmute(
      region = paste0(as.character(region), " (", n, ")"),
      `5 yr` = fixed_5y, `7 yr` = fixed_7y, `10 yr` = fixed_10y, expanding,
      guidance
    ) %>%
    knitr::kable(align = c("l", "r", "r", "r", "r", "l"))
}
```

**Default: expanding** — the only window viable everywhere it physically can be (94 / 96 countries eligible at some origin). **`fixed_10y`** carried as the trimming comparison for the data-rich regions (≈ expanding for coverage, diverges only where \>10 yr of history exists). **`fixed_7y`** as a Caribbean / Pacific fallback. **`fixed_5y`** diagnostic only — it cannot be a global default.
:::

::: {.callout-note title="Notes / observations"}
*Add: final window decision, how the chronically-intermittent countries (SSA, Caribbean tail) are handled, whether fixed-5yr is worth the compute as a diagnostic.*
:::

------------------------------------------------------------------------

# Model progress {#sec-progress}

Which models have a definition file, and which stages they've cleared so far.

```{r}
#| label: tbl-model-progress
#| tbl-cap: "Stage 0 status for every model with a Stage 0 record."

if (is.null(stage0_diag)) {
  pending("Scripts/forecasting/run_stage0_fit.R")
} else {
  stage0_diag %>%
    transmute(
      Model = model,
      `Run at` = run_at,
      `Stage 0` = if_else(pass, "PASS", "FAIL"),
      `Stage 1` = "-",
      `Stage 2` = "-",
      Notes = notes
    ) %>%
    knitr::kable()
}
```

```{r}
#| label: model-files-note
#| output: asis

model_files <- sort(list.files(
  file.path(repo_root, "Scripts", "forecasting", "models"),
  pattern = "^[0-9].*\\.R$"
))
have_record <- if (!is.null(stage0_diag)) nrow(stage0_diag) else 0L
cat(sprintf(
  "%d model definition file%s in `Scripts/forecasting/models/`: %s. %d have a Stage 0 record above; the rest are defined but not yet run (or parked).\n",
  length(model_files), if (length(model_files) == 1) "" else "s",
  paste0("`", model_files, "`", collapse = ", "),
  have_record
))
```

::: {.callout-note title="Notes / observations"}
*Add: which parked/unrun models need attention before Stage 1.*
:::

------------------------------------------------------------------------

# Stage 0 — model fit diagnostics {#sec-stage0}

`run_stage0_fit.R` fits each model once on the full panel and runs its diagnostic checklist (convergence, dispersion, residual structure). Diagnostic only — no skill claim.

```{r}
#| label: tbl-stage0
#| tbl-cap: "Stage 0 fit diagnostics."

if (is.null(stage0_diag)) {
  pending("Scripts/forecasting/run_stage0_fit.R")
} else {
  knitr::kable(stage0_diag)
}
```

::: {.callout-note title="Notes / observations"}
*Add: did every candidate pass the sanity check? Any coefficient signs or dispersion values worth flagging before Stage 1.*
:::

------------------------------------------------------------------------

# Stage 1 — hindcast skill {#sec-stage1}

Rolling-origin backtest across the four window lengths and the scenario strata. Primary metrics: WIS (log scale) and peak-timing difference. Secondary: MAE, RMSE, CRPS, PIT / coverage.

```{r}
#| label: sec-stage1-body

pending("the Stage 1 backtest runner (not built yet)")
```

Planned outputs to land here: relative-skill leaderboard by lead time, score table (model × lead × stratum, with `n_seasons`), score heatmap, per-country fan charts, PIT / coverage calibration.

::: {.callout-note title="Notes / observations"}
*Add as results arrive.*
:::

------------------------------------------------------------------------

# Conclusions & manuscript notes {#sec-conclusions}

Running synthesis for the forecasting paper (target: PLOS Comp Biol / PLOS NTD).

- **Data.** *...*
- **Strata.** *...*
- **Baselines.** *...*
- **Best model(s) so far.** *...*
- **Where the forecast is trustworthy / where it is not.** *...*

------------------------------------------------------------------------

# Open questions & directions {#sec-directions}

- Minimum complete-seasons floor below which a country defaults to the seasonal baseline (`[Q2.4]`).
- Spatial (region-grouped) blocking alongside the temporal windows (`§6`).
- Promotion rule: beat both baselines at *majority* vs *all* lead times (`§6`).
- Refit cadence for the hierarchical model — monthly vs quarterly (`[Q2.6]`).
- Operational-utility scoring design at \~83-country scale (`[Q5.2]`).
- Climate covariate sources and the GAM / hierarchical model structure.

------------------------------------------------------------------------

# Appendix: Model details {#sec-model-details}

One subsection per model, added as each clears Stage 0. Every subsection follows the same template — a forecast summary and an uncertainty-by-lead-time / region view — with room underneath for anything specific to that model.

::: {.callout-important appearance="simple"}
Everything here is a **Stage 0 snapshot**: one origin (the training panel's last month), no ground truth, produced by the `predict()` contract check in `run_stage0_fit.R` (`stage0_fit/<model>/wiring_check_forecasts.csv`). It shows what the model's output currently looks like — **not** how well it forecasts. Skill (does the interval actually cover the truth) has to wait for the Stage 1 backtest.
:::

```{r}
#| label: model-detail-helper

# Load one model's Stage 0 wiring-check forecasts, with region attached and a
# relative-interval-width column added. Shared by every model subsection
# below. NULL if that model hasn't been run through Stage 0 yet.
model_wiring_forecasts <- function(slug) {
  fc <- read_fc("stage0_fit", slug, "wiring_check_forecasts.csv")
  if (is.null(fc)) return(NULL)
  fc %>%
    mutate(
      origin_date = as.Date(origin_date),
      target_date = as.Date(target_date),
      # NA (not Inf) when .pred is 0, so it drops out of medians rather than
      # dominating them.
      rel_width90 = if_else(.pred > 0, (.pred_upper90 - .pred_lower90) / .pred, NA_real_)
    ) %>%
    left_join(region_lookup, by = "iso3")
}
```

## nowcast {#sec-model-nowcast}

```{r}
#| label: nowcast-load

nc_fc <- model_wiring_forecasts("nowcast")
```

```{r}
#| label: tbl-nowcast-summary
#| tbl-cap: "nowcast: Stage 0 forecast summary."

if (is.null(nc_fc)) {
  pending("Scripts/forecasting/run_stage0_fit.R nowcast")
} else {
  fmt <- function(x) formatC(x, big.mark = ",", format = "d")
  tibble::tibble(
    Metric = c("Origin", "Countries forecast", "No forecast (NA)",
               "Median .pred", "Min .pred", "Max .pred"),
    Value = c(
      format(unique(nc_fc$origin_date)),
      sprintf("%s / %s",
              fmt(n_distinct(nc_fc$iso3[!is.na(nc_fc$.pred)])),
              fmt(n_distinct(nc_fc$iso3))),
      sprintf("%s (%s)", fmt(sum(is.na(nc_fc$.pred))),
              scales::percent(mean(is.na(nc_fc$.pred)), 0.1)),
      fmt(median(nc_fc$.pred, na.rm = TRUE)),
      fmt(min(nc_fc$.pred, na.rm = TRUE)),
      fmt(max(nc_fc$.pred, na.rm = TRUE))
    )
  ) %>% knitr::kable(align = c("l", "r"))
}
```

```{r}
#| label: fig-nowcast-pred-by-horizon
#| fig-cap: "nowcast: distribution of the point forecast by lead time (log scale)."
#| fig-height: 3.6

if (!is.null(nc_fc)) {
  nc_fc %>%
    filter(!is.na(.pred)) %>%
    ggplot(aes(factor(horizon), .pred + 1)) +
    geom_boxplot(fill = "#4C72B0", alpha = 0.6, outlier.size = 0.8) +
    scale_y_log10(labels = scales::label_comma()) +
    labs(x = "Lead time (months)", y = ".pred + 1 (log scale)")
}
```

```{r}
#| label: fig-nowcast-uncertainty-lead
#| fig-cap: "nowcast: median relative 90% interval width by lead time. Width = (.pred_upper90 - .pred_lower90) / .pred."
#| fig-height: 3.6

if (!is.null(nc_fc)) {
  nc_fc %>%
    group_by(horizon) %>%
    summarise(median_rel_width = median(rel_width90, na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(horizon, median_rel_width)) +
    geom_col(fill = "#DD8452") +
    scale_x_continuous(breaks = sort(unique(nc_fc$horizon))) +
    labs(x = "Lead time (months)", y = "Median relative 90% width")
}
```

```{r}
#| label: fig-nowcast-uncertainty-region
#| fig-cap: "nowcast: median relative 90% interval width by region and lead time."
#| fig-height: 4

if (!is.null(nc_fc)) {
  nc_fc %>%
    mutate(region = factor(region, levels = region_levels)) %>%
    group_by(region, horizon) %>%
    summarise(median_rel_width = median(rel_width90, na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(horizon, region, fill = median_rel_width)) +
    geom_tile(colour = "white") +
    scale_fill_viridis_c(option = "magma", direction = -1, na.value = "grey90") +
    scale_x_continuous(breaks = sort(unique(nc_fc$horizon))) +
    labs(x = "Lead time (months)", y = NULL, fill = "Median\nrel. width")
}
```

### 2026 year-to-date and 6-month forecast, by region

Solid = observed 2026 (the panel's post-nowcast series); dashed = the Stage 0 forecast from the last observed month, with a light 90% band. One figure per region — a single panel with all 96 countries would be unreadable, and even within a region the y-axis is shared and linear, so very low-burden countries will sit flat near zero next to a country like Cuba or Brazil.

```{r}
#| label: nowcast-2026-forecast-data

if (!is.null(nc_fc)) {
  nc_origin <- unique(nc_fc$origin_date)

  # region is a factor on `panel` (set earlier) and plain character on `nc_fc`
  # (joined from country_region_lookup.csv) - cast both to character here so
  # bind_rows() below doesn't hit a factor/character type conflict.
  obs_2026 <- panel %>%
    filter(Year == 2026, !is.na(cases)) %>%
    transmute(iso3, region = as.character(region), date, value = cases,
              segment = "observed", lower90 = NA_real_, upper90 = NA_real_)

  # Anchor the forecast segment at the origin's observed value, so the dashed
  # line starts exactly where the solid line ends rather than jumping.
  fc_anchor <- panel %>%
    filter(date == nc_origin, !is.na(cases)) %>%
    transmute(iso3, region = as.character(region), date, value = cases,
              segment = "forecast", lower90 = cases, upper90 = cases)

  fc_future <- nc_fc %>%
    filter(!is.na(.pred)) %>%
    transmute(iso3, region = as.character(region), date = target_date,
              value = .pred, segment = "forecast",
              lower90 = .pred_lower90, upper90 = .pred_upper90)

  nc_2026_plot_df <- bind_rows(obs_2026, fc_anchor, fc_future) %>%
    mutate(region = factor(region, levels = region_levels)) %>%
    arrange(iso3, date)

  nc_2026_labels <- nc_2026_plot_df %>%
    filter(!is.na(value)) %>%
    group_by(iso3, region) %>%
    slice_max(date, n = 1, with_ties = FALSE) %>%
    ungroup()
}
```

```{r}
#| label: fig-nowcast-2026-by-region
#| output: asis
#| fig-cap: "nowcast: 2026 year-to-date and 6-month forecast, by region."

if (is.null(nc_fc)) {
  pending("Scripts/forecasting/run_stage0_fit.R nowcast")
} else {
  fig_dir <- tempfile("nowcast_2026_region_figs_")
  dir.create(fig_dir)

  for (rg in levels(nc_2026_plot_df$region)) {
    d <- nc_2026_plot_df %>% filter(region == rg)
    if (nrow(d) == 0) next
    lbl <- nc_2026_labels %>% filter(region == rg)
    n_ctry <- n_distinct(d$iso3)

    p <- ggplot(d, aes(date, value, group = iso3, colour = iso3)) +
      geom_ribbon(
        data = d %>% filter(segment == "forecast", !is.na(lower90)),
        aes(ymin = lower90, ymax = upper90, fill = iso3),
        colour = NA, alpha = 0.12, show.legend = FALSE
      ) +
      # Two layers, not one aes(linetype=) mapping: ggplot2 can't draw a
      # single line with both varying colour and a non-solid linetype, but a
      # fixed (non-aes) linetype per layer is fine.
      geom_line(
        data = d %>% filter(segment == "observed"),
        linewidth = 0.6, linetype = "solid"
      ) +
      geom_line(
        data = d %>% filter(segment == "forecast"),
        linewidth = 0.6, linetype = "22"
      ) +
      ggrepel::geom_text_repel(
        data = lbl, aes(label = iso3), size = 3, hjust = 0, direction = "y",
        nudge_x = 25, segment.size = 0.2, min.segment.length = 0,
        max.overlaps = Inf, show.legend = FALSE
      ) +
      scale_colour_viridis_d(option = "turbo") +
      scale_fill_viridis_d(option = "turbo") +
      scale_x_date(date_labels = "%b %Y", expand = expansion(mult = c(0.02, 0.2))) +
      scale_y_continuous(labels = scales::label_comma()) +
      labs(x = NULL, y = "Monthly cases")

    fname <- file.path(fig_dir, paste0(make.names(rg), ".png"))
    fig_h <- max(3.5, 0.3 * n_ctry + 2.2)
    ggsave(fname, p, width = 8, height = fig_h, dpi = 130)

    cat(sprintf(
      "\n\n#### %s\n\n<img src=\"%s\" style=\"width:100%%;\" alt=\"2026 observed and 6-month forecast for %s\">\n\n",
      rg, knitr::image_uri(fname), rg
    ))
  }
}
```

::: {.callout-note title="nowcast-specific notes"}
*Add: behaviour tied to this method specifically — e.g. next_season_method choice, calibration-file coverage, the \[UNDER REVIEW\] uncertainty redesign's expected effect on these plots once it lands.*
:::

::: {.callout-note title="Notes / observations"}
*Add.*
:::

------------------------------------------------------------------------

# Provenance {.unnumbered}

```{r}
#| label: provenance
#| echo: false

tibble::tibble(
  item = c("Data snapshot", "Forecast horizons", "Rolling windows (months)",
           "Min training months", "Origin start", "Rendered"),
  value = c(
    snapshot_date,
    paste(range(forecast_horizons), collapse = "-"),
    paste(names(roll_windows), unname(roll_windows), sep = "=", collapse = ", "),
    as.character(min_train_months),
    format(origin_start),
    format(Sys.time(), "%Y-%m-%d %H:%M")
  )
) %>% knitr::kable()
```

```{r}
#| label: sessioninfo
#| code-fold: true
#| code-summary: "R session"

sessionInfo()
```
