---
title: "V1 vs V2 Delay Correction Validation"
author: "DENV Global Observatory"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(knitr)
library(scales)

# rmarkdown sets wd to this file's directory — walk up to repo root.
find_project_root <- function() {
  if (dir.exists("Output") && dir.exists("Scripts")) {
    return(normalizePath("."))
  }
  start <- if (nzchar(knitr::current_input(dir = TRUE))) {
    knitr::current_input(dir = TRUE)
  } else {
    getwd()
  }
  dir <- start
  for (i in seq_len(10)) {
    if (dir.exists(file.path(dir, "Output")) &&
        dir.exists(file.path(dir, "Scripts"))) {
      return(normalizePath(dir))
    }
    parent <- dirname(dir)
    if (identical(parent, dir)) break
    dir <- parent
  }
  stop("Cannot find project root (expected Output/ and Scripts/ in an ancestor directory).")
}

project_root <- find_project_root()
setwd(project_root)

val_dir <- file.path(project_root, "Output", "validation", "V2_correction")

read_val <- function(fname) {
  path <- file.path(val_dir, fname)
  if (!file.exists(path)) {
    stop("Missing ", path, " — run 01_V2_correction_validation.R first.")
  }
  read_csv(path, show_col_types = FALSE)
}
```

# Overview

This report compares pipeline outputs from **V1** (PAHO-only `apply_reporting_correction`) and **V2** (`apply_delay_correction` for PAHO and WHO) for the same run date.

Run the analysis script first:

```r
source("Scripts/validation/V2_correction/01_V2_correction_validation.R")
```

Default snapshot directories (edit `run_date` in the R script to change):

- `Output/2026_06_11_V1/`
- `Output/2026_06_11_V2/`

# Run metadata

```{r metadata}
meta <- read_val("run_metadata.csv")
kable(meta)
```

# Files unchanged by correction

Season averages and historical monthly proportions are upstream of Step 4; row counts should match.

```{r unchanged}
kable(read_val("unchanged_files_summary.csv"))
```

# Backfill output (`DENV_cases_backfill_output.csv`)

```{r backfill-summary}
backfill_diffs <- read_val("backfill_case_diffs.csv")
source_switches <- read_val("backfill_source_switches.csv")

kable(
  meta %>%
    filter(dataset == "backfill") %>%
    select(n_v1, n_v2, n_joined, n_only_v1, n_only_v2,
           n_identical, n_case_diffs, n_source_switches,
           median_abs_delta, max_abs_delta)
)
```

## Case differences by source

```{r backfill-by-source}
kable(
  read_val("case_diffs_by_source.csv") %>%
    filter(dataset == "backfill")
)
```

## Source selection switches

Rows where the winning source differs for the same country–month (`r nrow(source_switches)` total).

```{r source-switches}
if (nrow(source_switches) > 0) {
  kable(source_switches)
} else {
  cat("No source switches.")
}
```

## Largest backfill deltas

```{r backfill-top}
kable(
  backfill_diffs %>%
    filter(!is.na(cases_v1), !is.na(cases_v2)) %>%
    arrange(desc(abs_delta)) %>%
    select(iso3, Year, Month, source_v1, source_v2, cases_v1, cases_v2, delta,
           correction_reason) %>%
    head(20)
)
```

```{r backfill-hist, fig.height=4}
backfill_diffs %>%
  filter(!is.na(cases_v1), !is.na(cases_v2), abs_delta > 0) %>%
  ggplot(aes(abs_delta)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Backfill |V2 − V1| case differences",
    x = "Absolute case difference",
    y = "Country–months"
  ) +
  theme_minimal()
```

# Nowcast output (`DENV_cases_nowcast_output.csv`)

Correction propagates through merge and nowcast; expect similar diff counts to backfill.

```{r nowcast-summary}
nowcast_diffs <- read_val("nowcast_case_diffs.csv")

kable(
  meta %>%
    filter(dataset == "nowcast") %>%
    select(n_v1, n_v2, n_joined, n_only_v1, n_only_v2,
           n_identical, n_case_diffs, n_source_switches,
           median_abs_delta, max_abs_delta)
)
```

```{r nowcast-by-source}
kable(
  read_val("case_diffs_by_source.csv") %>%
    filter(dataset == "nowcast")
)
```

```{r nowcast-top}
kable(
  nowcast_diffs %>%
    filter(!is.na(cases_v1), !is.na(cases_v2)) %>%
    arrange(desc(abs_delta)) %>%
    select(iso3, Year, Month, source, Data_status, cases_v1, cases_v2, delta) %>%
    head(15)
)
```

# Country tracking

Step-level inclusion flags; V2 WHO correction should increase `step_4b_who_after_correction` relative to V1.

```{r tracking-steps}
kable(read_val("country_tracking_step_summary.csv"))
```

```{r tracking-diffs}
diffs <- read_val("country_tracking_diffs.csv")
if (nrow(diffs) > 0) {
  kable(
    diffs %>%
      select(iso3, country, step, value_v1, value_v2,
             final_status_v1, final_status_v2) %>%
      head(30)
  )
} else {
  cat("No country-tracking differences.")
}
```

Pipeline funnel summaries:

```{r tracking-summary}
kable(read_val("country_tracking_summary_v1.csv"))
kable(read_val("country_tracking_summary_v2.csv"))
```

# V2 correction audit

Summary of PAHO weekly and WHO monthly correction from `inital_rf_correction/` (V2 run only).

```{r audit}
audit <- read_val("correction_audit_summary.csv")
if (nrow(audit) > 0) {
  kable(audit)
} else {
  cat("No audit files found under V2 output.")
}
```

# Belize (BLZ) spot-check

Belize is excluded from PAHO correction in V2. Applied cases should equal raw; V1 and V2 case counts should match.

```{r belize}
belize <- read_val("belize_backfill_check.csv")
kable(
  belize %>%
    summarise(
      n_rows = n(),
      n_v1_eq_v2 = sum(belize_expected, na.rm = TRUE),
      n_applied_eq_raw = sum(cases_v2 == raw_cases, na.rm = TRUE),
      n_corrected_na = sum(is.na(corrected_cases), na.rm = TRUE)
    )
)
```

# Countries with largest total impact

```{r country-impact}
kable(
  read_val("case_diffs_by_country.csv") %>%
    filter(dataset == "backfill") %>%
    head(15)
)
```
